% !TEX root = dissertation.tex

\nocite{dillamnn_jbw_kleine}
\nocite{ewald_zkm1844}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEMORY CONSTRUCTION is the key Idea for this chapter. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
% TODO: Introduction to the chapter as a whole

% ``Constructing Authority: Restructuring and reinforcing the Past''
% Begin with the exercise of authority and the way that Jubilees reappropriates Moses
% Show how that authority is wielded through the imposition of the calendar and chronological system
% use of *written* things to reinforce practice

\subsection{Discovery and Publication}
% Discuss "discovery" 
The work we now referred to as the book of \jub was believed to have been lost forever by European scholars prior to the mid 19th Century. The work was ``rediscovered,'' however, in 1844 when Heinrich Ewald published a description of an Ethiopian (\geez) manuscript which had under the title ``the Book of the Division'' \eth{መጽሐፈ ኩፋሌ}{masḥafa kufālē}.%
        \footnote{All translations are my own. \geez citations are from \vanderkam's critical edition, \cite*{vanderkam1989}.}
Because the name followed the common convention using a work's first few (key) words as its title (in this case, \eth{ዝንቱ፡ነገረ፡ኩፋሌ}{zentu nabara kufālē}), Ewald suggested that this manuscript may have been a copy of the work known from antiquity as both \greek{τά Ἰωβηλαϊα}, ``the Jubilee,'' and \greek{Λεπτὴ Γίνεσις}, the ``Little Genesis.''\autocite[176--179]{ewald_zkm1844} Although the work had been in continuous use within Ethiopian Christianity since antiquity, European scholarship only knew of the work through secondary references in a few classical sources.%
        \footnote{\vanderkam offers a concise summary of the various late-antique citations and allusions in his commentary, most notably in the works of Epiphanius (\emph{Panarion}, \emph{Measures and Weights}) and Syncellus (\emph{Chronography}).
                \cite[1:10--14]{vanderkam2018}. See also 
                \cite{reed_kister-etal2015} and 
                \cite{kreps_ch2018}.
        It is also probable that more recently discovered text, such as the Damascus Document (CD refer to the book of \jub as 
        ``the Book of the Divisions of the Times into their Jubilees and Weeks'' Heb. \hebrew{ספר מחלקות העתים ליובליהם ובשבועותיהם}. It seems almost unimaginable that CD was not refering to \jub, though, some have questioned the notion. See \cite[242--248]{dimant_vanderkam-etal2006}.}

The work was published and supplemented by an additional manuscript by August Dillmann in 1859\autocite{dillmann1859} and R.~H. Charles in 1895, who included two additional manuscripts in his edition (totaling four).\autocite{charles1895} More recently, \vanderkam's 1989 edition utilized twenty-seven copies of the text\autocite[1:xiv--xvi]{vanderkam1989} and since its publication over twenty more copies have been cataloged and imaged.%
        \footnote{%
                \cite{erho_bsoas2013}.
                \vanderkam helpfully lists the twenty-seven manuscripts he used for his critical edition in the introduction of his commentary where he also notes the additional manuscripts photographed since its publication. See 
                \cite[1:14--16]{vanderkam2018}.}

With the exception of the ``rediscovery'' of the text for European scholarship, the most significant find for the study of \jub was the discovery of several Hebrew fragments among the \dss. These fragments  attest to the work's antiquity and confirmed that the original language of \jub was Hebrew and not Aramaic, as Dillman had supposed.%
        \footnote{TODO:}
Ethiopic text is to be a granddaughter translation of the Hebrew through Greek, though Greek manuscripts of the text have been found.\footnote{See especially \vanderkam's treatment of the textual history of \jub in \cite*[1--18]{vanderkam1977}.} This fact was convincingly demonstrated by Dillmann who observed several Greek forms preserved as transliterations in the Ethiopic text.\footnote{Specifically: \greek{δρῦς}, \greek{βάλανος}, \greek{λίψ}, \greek{σχῖνος}, and \greek{φάραγξ}. \cite[88]{dillamnn_jbw1850}. Charles later added \greek{ἡλιου} to the list. \cite[xxx]{charles1902}.} By the end of the 19th century, however, partial copies of \jub had also been uncovered in Latin, which similarly appear to have come through the Greek. Finally, although no direct manuscript evidence has been found, \jub scholars posit that a Syriac translation of the Hebrew was made in antiquity. This suggestion tenuous, but is based on a number of Syriac citations of \jub which do not show any linguistic influence (loan words, etc.) from Greek.%
        \footnote{%
                See especially
                \cite[231--232]{tisserant_rb1921} and 
                \cite[xxix]{charles1902} but also 
                \cite[2:ix--x]{ceriani1861} and 
                \cite[x]{charles1895}.}
Despite all of these finds, however, the Ethiopic text remains the only tradition to preserve \jub in its entirety. Thus, in my treatment of \jub I will be relying primarily on the Ethiopic text and will be supplementing from the Hebrew where available.

\subsection{Content and Character}
The book of \jub offers a rewriting of the book of Genesis and the first part of Exodus (Gen 1--Exod 12).\autocite[1:17]{vanderkam2018}  The prologue gives a short description of the work as an account concerned with the division of time into units of years, weeks, and jubilees given to Moses when he ascended Mt. Sinai to receive the ``stone tablets'':

\input{includes/jubilees_prologue}%
\noindent
The bulk of the work (\jub 2:1--50:13) is dedicated to the recounting of the events of Gen 1--Exod 12 with special concern for halakhic matters and the division of time according to ``weeks'' of years (7-year units) and ``jubilees'' (49-year units). The particulars of the revelation are mediated by the ``\ap'' (Eth. \ethiopic{መልአከ ገጽ} [\translit{mal'aka gaṣṣ}]) who dictates the content of the ``heavenly tablets'' (1:27; Eth. \ethiopic{ጽላተ፡ሰማይ} [\translit{ṣəllāta samāy}]) to Moses to record what they revealed about the structure and terminus of the cosmos.\autocite{martinez_najman-tigchelaar2012} The treatment of Moses as a scribe places him within a chain of tradition---along with Enoch and Noah---which emphasizes writing and written works as essential sources of tradition and revelation.%
        \footnote{See especially
                \cite[381--388]{najman_jsj1999}.} 
The work closes with a terse statement declaring ``Here the account of the division of time is ended'' (\jub 50:13; Eth. 
    \eth{ተፈጸመ ፡ በዝየ ፡ ነገር ፡ ዘኩፋሌ ፡ መዋዕል ።}
        {tafaṣṣama ba-zeyya nagar za-kufālē mawāʕel}).
        % tafaṣṣama     tD fṣm 'to complete'
        % ba-zeyya      here
        % nagar         account, speech, etc.
        % kufālē        division(s)
        % mawāʕel       √mʕl 'to pass the day' here: period, era, time Les. 603
Thus the main body of the work is presented as a revelation from \yahweh given to Moses atop Mt. Sinai, framed by a brief prologue and epilogue which situates the story during Moses's first 40-days atop the mountain when Moses first receives the Tablets of Stone (Exod 24:12).\autocite[1:129]{vanderkam2018}


\subsection{As RwB}


\subsection{Thesis on Memory}
% MEMORY CONSTRUCTION is the key Idea for this chapter. 
%% Calendar is the schema for a new conception of the past
%% Authority conferring strategies as memory-making strategies
%% Writing and language as reinforcement of tradition
