% !TeX root = ../dissertation.tex

\chapter{Narrative, Genre, and \Psy: \GA as Cultural Memory}
\label{chap:ga}

Since its initial discovery and publication, the Aramaic text known as \ga (\q{1}{apGen ar}{}) has been associated in various ways with the book of Genesis. As one of the first seven scrolls discovered in the Judean desert beginning in 1947, \ga is also one of the better studied works among the Dead Sea Scrolls. When the scroll was initially analyzed by scholars, it could not be fully unrolled and only a small portion of the outer layer of the scroll could be read. These visible portions, written in Aramaic, referenced the antedeluvian Lamech, the father of Noah, and his wife, Bitenosh, known from the book of \jub. The text appeared to be written in as a first-person narrative from the perspective of Lamech, leading Trevor to conclude that the scroll was a copy of the so-called ``Book of Lamech'' listed as an apocryphal work by a \ce{seventh-century} Greek canon list.%
    \footnote{This fact led Trevor to refer to the scroll as the ``Ain Feshkha Lamech Scroll'' and Milik to refer to it as the ``Apocalypse of Lamech'' for the publication of the fragment in DJD 1. See \cite[9--10]{trevor_basor1949} and ``Apocalypse de Lamech'' in \cite[86--87]{djd_1}.}
Once the scroll was completely unrolled, it became obvious that the scroll contained more than just a first-person account from Lamech and instead contained additional first-person accounts from figures found in the Genesis stories, including Noah and Abram. Thus, the more descriptive title, \citetitle{avigad-yadin1956}, was given to the scroll by Avigad and Yadin for the publication of its \emph{editio princeps} in 1956.%
    \footnote{Hebrew \he{מגילה חיצונית לבראשית}. See \cite{avigad-yadin1956}.}
While the name \ga remains in wide use, it is notable that the name has been criticized and a number of alternative titles have been suggested; most notably: ``Book of the Patriarchs,''%
    \footnote{Hebrew \he{ספר אבות}. As suggested by Mazar in \cite[379 n. 2]{flusser_ks1956}.}
``Memoirs of the Patriarchs,''%
    \footnote{\Cite[358]{gaster1976}.}
and \aram{כתב אבהן}.%
    \footnote{\Cite[14 n. 1.]{milik1959}. Fitzmyer suggests \aram{כתב אבהתא} would be, perhaps, even more suitable. See \cite[16]{fitzmyer2004}.}
In this chapter, I will retain the traditional title, \ga.

Although much of the scroll was very badly damaged, illegible, or missing, enough survived for Avigad and Yadin to make the generalized observations that \ga followed the basic order and events of Genesis from the Flood (Gen 6) into the Abram narrative (ending in Gen 15). The events are generally (though not exclusively) narrated in a series of first person accounts---what I will refer to as ``memoirs''%
    \footnote{I will use the term ``memoir'' throughout this chapter as a way of referring to the distinct (mostly) first-person narratives found in \ga. The term is meant to highlight the formal characteristic of being written in the first person voice without any reference to the authenticity of the work and in alignment with the convention of referring to first-person narratives in the Bible as ``memoirs'' (e.g., the ``Nehemiah Memoir'' or the ``Isaiah Memoir''). For a discussion of characterizing the structure and nature and genre of \ga see \cite{bernstein_berthelot-etal2010} and \cite{bernasconi_as2011}.}%
---by Lamech, Noah, and Abram, respectively and show a clear affinity with the roughly contemporaneous works of \firstenoch and \jub.%
    \autocite[16--37]{avigad-yadin1956}
The literary relationship of \ga to both \firstenoch and (especially) \jub remains a matter of debate, with Avigad and Yadin suggesting that \ga more probably preceded \jub, while the recent prevailing opinion seems to prefer the opposite.%
    \footnote{\Cite[38]{avigad-yadin1956}; cf. \cite[20--21]{fitzmyer2004}. Fitzmyer cites Hartman's suggestion, building on Fitzmyer's own work, treating the similarity between \ga's and Jubilee's chronology of Abram's life. Because the chronology seems to have been closely tied to Jubilee's uniquely structured calendar, it follows that \ga drew from \jub. See \cite[497]{hartman_cbq1966}.}

The name given to \ga in the \emph{editio princeps} set the agenda for scholarly inquiry on the work into the modern era by connecting it to the biblical book of Genesis while simultaneously categorizing it as apocryphal. Much of the attention given to \ga, therefore, has focused on its literary genre and its relationship to the Bible and resemblance of the Targums and later midrashic works. As already noted, \vermes's treatment of \ga focused on the role that it played in showing the continuity between the interpretation of Jewish scripture during the \secondtemple period and the aggadic traditions of early rabbinic Judaism. In \emph{Scripture and Tradition}, \vermes treats in detail the relationship between Gen 12:8--15:4 and \ga \cols{19}{22}, ultimately declaring \ga to be ``the most ancient midrash of all'' and the ``lost link between the biblical and the rabbinic midrash.''%
    \autocite[124]{vermes1961}
The result of this framing (whether one considers it appropriate or not) has been that much of the scholarly attention paid to \ga has focused on its relationship to Genesis and especially how its author(s) may have been addressing exegetical issues found within the biblical work. Yet, as Fitzmyer observes, the roots of biblical midrash are now generally accepted to be found within the Hebrew Bible itself.%
    \autocite[20]{fitzmyer2004}
The fact that a number of Targums have been found at Qumran makes the presence of targumic and midrashic qualities in \ga less remarkable and frees us from any obligation to try and fit it cleanly within either category.

Although few scholars insist on rigidly defining \ga as either targum or midrash, the treatment of \ga as primarily \emph{exegetical} tacitly implies that the purpose of \ga was to explain or interpret Genesis. Put another way, the discussions surrounding \ga are often preoccupied with gleaning information about how \secondtemple Jews read \emph{Genesis}---treating \ga from the perspective of ``biblical interpretation.'' While there is no question that such an approach has been fruitful, treating \ga as \emph{only} or even \emph{primarily} an example of biblical interpretation, I think, cannot offer a comprehensive reading of the work. In this chapter, therefore, I will focus on the ways that the author of \ga engages with a  constellation of discourses surrounding events and characters \emph{known from} the book of Genesis, as well as those from other texts not as biblical \emph{interpretation}, but as processes of \emph{memory}.

What remains uncertain about \ga is what its function may have been for its original audience. I am in agreement with Fitzmyer that it seems unlikely that \ga would have been used liturgically and that the general character of the work is ``for a pious and edifying purpose.''%
    \autocite[20]{fitzmyer2004}.
Yet, I can not help but feel somewhat dissatisfied with this answer. How might \ga have edified its readers? Works such as \jub and \firstenoch, perhaps, have more obvious rhetorical aims, but for all its similarities to these texts, \ga maintains a different character which has generally eluded commentators. While I have no illusions that I will be able to offer a satisfactory answer to the question of \ga's specific purpose, approaching \ga as an object of cultural memory, I believe, is a good place to start. The advantage that a memory approach has in addressing this problem is that it offers a way to talk about the manifold ways that \ga both builds from its cultural memory and speaks back into it.

From the perspective of cultural memory, therefore, \ga operates within a stream of traditions and participates in discourses surrounding early foundational figures in Jewish tradition: Lamech, Noah, and Abram. As such, it is both the recipient and progenitor of cultural memory whose participation in the mnemonic process affected the memory itself. Thus, building on the theoretical framework of \hyperref[chap:memory]{chapter two}, we may understand \ga to have taken part in three discrete mnemonic processes: 1) the reception of cultural memory, 2) the reshaping of memory by contemporary social frameworks (social memory), and 3) the active construction, codification, and reintegration of memory for future transmission. In this chapter, therefore, I have chosen to frame the discussion of \ga as \rwb with reference to these mnemonic processes. First, and as a point of departure, I will discuss the ways that \ga functions as the recipient of cultural memory through its engagement with what I refer to as  ``biblical memory.'' Second, I will discuss the ways that \ga was shaped by the social frameworks which inherited it, through a discussion of literary genre and shared formal characteristics with contemporary texts. Finally, I will discuss how \ga participated in the construction of cultural memory through its use of \psgraphical discourse.


\section{\ga and Biblical Memory}

Although it is anachronistic to suggest that the ``Bible''  existed during the late \secondtemple period, the ``biblical'' texts found at Qumran provide evidence to support the idea that many of the texts and traditions that were later formalized as the ``Bible'' were present in a reasonably stable and even privileged state near the turn of the era. In other words, while I wish to push the discussion away from what I see as a preoccupation with the relationship between \rwb texts their putative biblical counterparts, I do not mean to deny the reality that certain ``biblical'' texts certainly held uniquely privileged positions within the religious and social milieu of \secondtemple Judaism. Therefore, I think it is a mistake to jettison any discussion of \rwb texts as they relate to the texts that would later become the Hebrew Bible. On the other hand, restricting our discussion to those later biblical texts would likewise not do justice to the wide variety of texts and traditions in existence during the \secondtemple period which undoubtedly influenced \ga. In an effort to strike a middle ground, therefore, I have opted to refer to \ga as it relates to ``biblical memory,'' a term which deserves some unpacking. By biblical memory, I have in mind the set of discourses that engage with sites of memory known to us through the biblical text; the stories and traditions---irrespective of any notion of authority or canon---which make up the expansive network traditions related to the text of the Hebrew Bible.%
    \footnote{I would like to emphasize that I am not suggesting that ``biblical memory'' represents a qualitatively unique form of memory, only that the scope of the traditions under consideration relate to texts that later became the Bible, and, in all likelihood, held at least some sort of special privilege within the memory of many \secondtemple Jews. In other words, by ``biblical memory'' I mean a strict subset of cultural memory which happen to relate to the Hebrew Bible.}
This rhetorical move is meant to blur the line between the \rwb text's putative biblical ``\emph{Vorlage}'' and the broader cultural perception---both official and popular---of the events and characters with which the \rwb text deals. In this section, therefore, I will discuss the ways that \ga participated in biblical memory through a discussion of its relationship to both the biblical text and other \secondtemple texts such as \jub and \firstenoch.

% What was GA rewriting: Sources for the parts of GS
Although \ga is generally touted as one of the more clear-cut examples of the \rwb, it is noteworthy that its relationship to the biblical text is not, in fact, entirely uniform.%
    \autocite[333]{bernstein_berthelot-etal2010}

% First Columns (Lamech Memoir: Summary and relation to 1 Enoch)
\subsection{The Lamech Memoir (Columns 0--\rom{5})}
The earliest columns of \ga (cols. 0--\rom{5}), which are narrated from the perspective of Lamech (the ``Lamech Memoir'' by my terminology), Noah's father, essentially offer a rewriting of \firstenoch 106--107.%
    \footnote{\Cite[174]{nickelsburg2005}. The birth of Noah seems to have been a matter of some interest; a number of other texts likewise discuss the exceptional qualities of Noah at his birth. See \q{4}{534}{}[\q{4}{BNoah}{a-d}], \q{1}{Noah} as well as \cite{vanderkam_kapera1992}. On the first-person narration of this section, see \cite{weigold_as2010}. Note also \cite{stuckenbruck_berthelot-etal2010} and \cite{baxter_jsp2006}.}
In this section, Lamech, recounts the birth of Noah and Lamech's fear that his wife, Bitenosh, had conceived Noah by means of the \aram{עירין} ``Watchers.'' Despite Bitenosh's assurances, Lamech petitions his father, Methuselah, to ask \emph{his} father, \enoch, for further assurance, which he ultimately gives. Although this section is fragmentary, its close resemblance to \firstenoch 106--107 makes the scholarly reconstruction of the missing sections quite plausible. While it may be tempting to suggest that this section of \ga represents a variant edition of \firstenoch 106--107, rather than a rewriting, the fact that the version of the story preserved in \ga is told in the first person from the point of view of Lamech, while \firstenoch 106--107 is told in the third person, makes this suggestion highly unlikely. Moreover, because both \firstenoch and \ga were composed in Aramaic, the differences between the two tellings cannot be attributed to translational issues. In other words, although cols. 0--\rom{5} deal, nominally, with events in Genesis 5:28--29, for all intents and purposes, the story recounted in these columns is a retelling of events known from the enochic tradition and \emph{not} the book of Genesis.%
    \footnote{It is not clear what the precise relationship between the enochic traditions and \ga actually was. Here I have essentially assumed the priority of \firstenoch, but I wish to leave ambiguous whether \ga represents a rewriting of the \emph{text} of \firstenoch, or whether they simply draw on a common tradition. Thus, I have chosen to refer to the tradition ``known from'' \firstenoch, rather than \firstenoch itself. See Stuckenbruck's treatment of these traditions in \cite*{stuckenbruck_berthelot-etal2010}; Nickelsburg's concise but thorough treatment of the similarities and differences in of these texts is also quite helpful. See \cite[173--174]{nickelsburg2005} as well as \cite[122--123]{fitzmyer2004}.} 


% Second Section (Noah Memoir)
\subsection{The Noah Memoir (\Cols{5}{17})}
The second major section of \ga begins with a superscription identifying what follows as a \aram{[פרשגן] כתב מלי נוח} or ``[A copy of] the Book of the Words of Noah'' (\col{5}, 29) and continues through \col{17} (and, likely, onto the beginning of \col{18}).%
    \footnote{\Cite[174--175]{nickelsburg2005};
        Regarding the superscription, see \cite{steiner_dsd1995}.
        On the topic of the existence of a so-called ``book of Noah'' see 
        \cite{dimant_vanderkam-etal2006} and \cite{werman_chazon-etal1999}.}

% Summary
Although this section accounts for the bulk of the scroll, significant portions are missing or unreadable. This ``Noah Memoir'' begins with a description of Noah's righteousness%
    \autocite{vanderkam_collins-nickelsburg1980}
(affirmed even in utero) and his early family life (\col{5}, 29--\rom{6}, 9), followed by a vision predicting the flood (\col{6}, 9--\rom{7}, 9) which comes about due to the evil behavior of the Nephilim. \Cols{7}{8} are highly fragmentary, but most likely described the events of the flood, while \cols{9}{12} (which are slightly less fragmentary) describe the ark's putting in on Mt. Ararat, God's instructions to  and blessing of Noah (including the prohibition of consuming blood), and Noah's subsequent interest in viticulture. \Cols{13}{15} recount a dream-vision in which Noah is depicted as a cedar tree with shoots representing his sons, including a fragmentary explanation of the dream. Finally, \cols{16}{17} describe the division of the land by Noah to his sons.

% ``Sources'' and Relationship of NM to Jubilees & 1 Enoch?
As with the Lamech Memoir, the Noah Memoir clearly draws from traditions outside of those preserved in Genesis. This fact was acknowledged even from the scroll's initial publication.%
    \autocite[38]{avigad-yadin1956}
Although the flood account in Gen 6:9--9:17 is a longer and more developed story in its own right than is the account of Noah's birth (which the Lamech Memoir takes as its point of departure), characterizing either cols. 0--\rom{5} or \cols{6}{17} of \ga as \emph{primarily} a rewriting of Genesis does not give due consideration to the additional traditions which influenced its composition. The extended treatment of Lamech's reaction to Noah's birth in cols. 0--\rom{5}, including the mention of Lamech's wife Bitenosh, betrays the scroll's extra-biblical sources (esp. \firstenoch and \jub, respectively; more on this below). The mention of the Watchers (Aramaic \aram{עירין}) and the Nephilim in \cols{6}{7} also evinces a clear thematic affinity to the Book of Watchers in \firstenoch 6--11.%
    \footnote{\Cite[174]{nickelsburg2005}.}
Moreover, the explicit reference to the ``the [Book] of the Words of Enoch'' in \col{19}, 25 suggests that \ga was familiar with \firstenoch, or at the very least a tradition of enochic writings.%
    \footnote{It is worth noting that this reference occurs in the latter Abram section, which some have argued originates in a different source than the first two memoirs. See esp. \cite{bernstein_berthelot-etal2010} and \cite{bernstein_as2010}.}

Especially plain, however, is the Noah Memoir's connection to the book of \jub, which seems to offer a consistent point of contact with this section of \ga.%
    \autocite[20]{fitzmyer2004}
In fact, it was the explicit identification of Lamech's wife Bitenosh which first prompted Trevor's initial identification of the (unopened) scroll with the so-called Book of Lamech.%
    \autocite{trevor_basor1949}
Although an exhaustive treatment of the parallels between \jub and \ga is outside the scope of this chapter, it will suffice to note a few of the most significant points of contact between the Lamech and Noah Memoirs and \jub. James \vanderkam has recently offered a detailed, yet concise, summary of these similarities and differences, which, while too long to reproduced in full, can be summarized as follows:%
    \footnote{See \cite[374--376]{vanderkam_feldman-etal2017}. For additional treatments of this topic, see also 
    \cite{machiela2009}; \cite{machiela_mason-etal2012} and 
    \cite[305--342]{kugel2012}, previously published as 
    \cite{kugel_roitman-etal2011}.} 

\begin{quote}
\begin{enumerate}
    \item Several personal and geographic%
            \footnote{Mahaq Sea (\col{16}, 9; \jub 8.22), Tina River (\col{16}, 15; \jub 8.12), Mount Lubar (\col{12}, 13; \jub 5.28), Erythrean/Red Sea (\col{17}, 7; \jub 8.21), and Gadeira (\col{16}, 11; \jub 8.26).}
        names which are never mentioned in the Bible show up in both \ga and \jub (including Bitenosh, which is a part of the Lamech Memoir).
    \item Both \jub and \ga utilize ``\jub'' as a significant chronological unit (\ga to a lesser degree than \jub).
    \item Several shared stories, themes, and phrases such as 1) ``in the days of Jared,'' 2) Enoch remaining accessible after his departure from normal terrestrial life, 3) Noah making atonement for the ``whole earth,'' and 4) stories recounting Noah and his vineyard.
    \item The ``division of the earth,'' while different in several specifics, are strikingly similar and offer, perhaps, the most compelling case for a direct, genetic relationship between the two texts.%
        \footnote{See also Machiela's extensive treatment of this section where he argues for the theory that both texts could be drawing from a shared cartographical source in \cite*[105--130]{machiela2009}. See also \cite{alexander_jjs1982}.}
\end{enumerate}
\end{quote}

The striking similarities between the early columns of \ga and \jub (and to a lesser degree, \firstenoch) over and against the biblical text, complicates the characterization of \ga as \rwb or strictly exegetical in nature. In other words, if \ga drew from \jub (or if they drew from some common source) I think it is fair to scrutinize whether this section of \ga should be considered a rewriting of \emph{Genesis}. What is clear, instead, is that for the author of \ga, the memory of the flood (and adjacent characters and events) did not center on the biblical account. The question could, therefore, be asked whether cols. 0--\rom{17}---the bulk of the scroll---would meet the generic criteria of \rwb according to the definitions of Bernstein and other more conservative commentators. My guess is that if \cols{19}{22} had also been lost, \ga would more commonly be categorized as ``parabiblical'' like \firstenoch.%
    \footnote{Notably, this is the preferred nomenclature of Fitzmyer, though he also considers \ga to be ``a good example of the so-called rewritten Bible.'' \cite[20]{fitzmyer2004}.}

% Third Section (Abram) Description
\subsection{The Abram Memoir (\Cols{19}{22})}

The final surviving columns of the scroll, \cols{19}{22}, represent the longest and most complete sustained narrative preserved in \ga, here referred to as the ``Abram Memoir.''%
    \footnote{On the ``memory'' of Abram more broadly, see \cite{ego_falk-etal2010}, \cite{benzvi_edelman-benzvi2013}, \cite{hendel2005} and \cite{reed_jsj2009}.}
More so than the previous sections, the Abram Memoir maps very closely onto the events narrated in Genesis.%
    \footnote{On the particular text tradition that \ga builds from, see \cite{vanderkam1978}.}
These columns parallel Genesis 12:10--15:14, retelling the stories of Abram and Sarai's sojourn in Egypt (|| Gen 12:10--20), Abram's subsequent conflict with Lot (|| Gen 13:1--18), the Elamite campaign (|| Gen 14:1--24), and the beginning of Abram's vision (|| Gen 15:1--4). \GA's retelling of these stories follows the chronology of Gen 12--15 very closely, but embellishes and augments the narrative throughout. Like the Lamech and Noah Memoirs, this section of \ga is largely written as a first-person narrative, this time in Abram's voice. The transition between the Noah Memoir and the Abram memoir is missing, so there is no superscription or title for this section. Throughout this section, however, the phrase ``I, Abram'' shows up a number of times, making it clear who the narrator is. Yet, this fact is complicated by the fact that beginning at \col{21}, 23 the narrator transitions to the third person and remains in this voice through the end of the surviving portion of the scroll.%
    \footnote{It is worth pointing out that the final surviving sheet of parchment was not the final sheet of the scroll originally. Avigad and Yadin note that although only four sheets of the work were present, the seam between the fourth and (what would be) the fifth sheets is visible on the edge of the fourth sheet. \cite*[14]{avigad-yadin1956}.}
This inconsistency, perhaps more than any other feature of \ga, has complicated its generic classification.

% First Half (Midrash)
The earlier portions of the Abram Memoir strike a balance between fidelity and innovation with regard to the \emph{biblical} text that the other sections lack. For example, the narrative of Abram and Sarai's descent into Egypt is clearly and recognizably built from the story preserved in the Hebrew Bible. The events and chronology of the story map directly onto Gen 12:10--20, but \ga offers---in addition to the first-person point of view---a number of expansions that seem plainly to be innovative or, as \vermes would put it, an example or prototype of ``midrash.''%
    \footnote{\Cite[124]{vermes1961}. Notably, the characterization of \ga as \rwb is typically based on an analysis of the Abram Memoir. Although the earlier portions of the scroll were known, \vermes's treatment of \ga only dealt with \cols{19}{22}. Together with the fact that these are the best-preserved and most complete columns, this fact has, I think, impacted the characterization of \ga as a whole, perhaps unfairly. On the characterization pre-rabbinic texts as ``midrash,'' see \cite[esp. 298--305]{mandel2017}; \cite{mandel_dsd2001}; \cite{mandel_bakhos2006}.}
Numerous small additions and emendations occur throughout the retelling, such as making explicit how long Sarai and Abram lived in Egypt prior to Sarai's notice by Pharoah's princes, how long Sarai was with Pharoah, and numerous geographical and personal names. A number of these details, as with earlier sections of \ga, are also found in \jub, underscoring the close (if poorly understood) relationship between the two texts.

More noticeable are the larger expansions present in \ga such as Abram's portentous dream (\col{19}, 14--17), the \emph{waṣf} put on the lips of Pharoah's princes about Sarai (\col{20}, 2--8),%
    \autocite{vanderkam_revq1979}
Abram's prayer following Sarai's abduction (\col{20}, 12--16), the details of Pharoah's afflictions (\col{20}, 16--21), Harkenosh's discussion with Lot (\col{20}, 21--\rom{20}, 24), and Abram's intervention on Pharoah's behalf (\col{20}, 24--32).%
    \footnote{Other changes from later in the memoir include a description of Abram walking the length and width of the land as well as a notable abbreviation of Abram and Lot's conflict in Gen 13:5--12.}

The explanation of these expansions, according to \vermes---whose view has been adopted by most treatments of \ga---is as a means of ``correcting'' or otherwise supplementing the biblical text in order to engage the reader and to \emph{explain} the biblical text.%
    \autocite[126]{vermes1961}
\vermes writes:

\begin{quote}
    The author of GA does indeed try, by every means at his disposal, to make the biblical story more attractive, more real, more edifying, and above all more intelligible. Geographic data are inserted to complete biblical lacunae or to identify altered  place names, and various descriptive touches are added to give the story substance\dots. To this work of expansion and development Genesis Apocryphon adds another, namely, the reconciliation of unexplained or apparently conflicting statements in the biblical text in order to allay doubt and worry.\autocite[125]{vermes1961}
\end{quote}
% Latter Half (Targum)
\noindent
By contrast, the latter portion of the Abram Memoir, beginning at \col{21}, 23, at times borders on a word-for-word translation of Genesis into Aramaic with comparatively few significant changes. This quality provided occasion for a number of (especially early) scholars to compare \ga with the Targums.%
    \footnote{\Cite[193]{black1983}. Though he notably amended his opinion later in \cite*{black_black-fohrer1968}.}
Although the change from first person to third person is, perhaps, the most significant literary shift that occurs in \ga, other literary features of the Abram Memoir agree against the Lamech and Noah Memoirs in such a way that they give reason to suppose the Abram Memoir makes up a literary unit.%
    \footnote{Specifically, Moshe Bernstein has noted based on the divine names that are used throughout the work that the primary division is between the Lamech/Noah Memoirs and the Abram Memoir; the earlier sections utilizing a specific set of divine titles and the latter section(s) using a different set. See \cite{bernstein_jbl2009}; See also \cite[97]{falk2007}. Regarding the genre(s) and unity of \ga more generally see Bernstein's later work \cite*{bernstein_berthelot-etal2010} and \cite*{bernstein_as2010}.}
It is not clear, however, why there seems to be such a dramatic difference in narrative voice beginning at \col{21}, 23.

\subsection{Exegesis and Memory}
Emphasizing the supposed ``exegetical'' quality of \ga, some modern treatments of the work still characterize \ga as a form of Targum or Midrash, with a preference for the latter.%
    \footnote{\Cite{evans_revq1988}; \cite[19]{fitzmyer2004}. Esther Eshel has proposed the term ``narrative midrash,'' but I am in agreement with Harrington and Bernstein in eschewing later categories such as ``midrash'' for these pre-rabbinic sources. See \cite[182]{eshel_roitman-etal2011}; Cf. \cite[242]{harrington_kraft-nickelsburg1986}; \cite[327 n. 33; 328--329]{bernstein_berthelot-etal2010}. On the particular ``methods'' used by the author of \ga, see \cite{bernstein_dsd1996} and \cite{bernstein_henze2012}.}
Yet, as I have illustrated, although portions of \ga relate clearly to the text of Genesis (notably, the Abram Memoir), much of the earlier portions of the scroll only nominally relate to Genesis, and instead show an affinity to the traditions associated with \firstenoch and \jub. Thus, characterizing the work as focused primarily on the explanation of Genesis (as \vermes suggests), is ill-founded, by my reckoning. Indeed, the disjunction between the various parts of \ga have been observed by numerous scholars, even by those who broadly accept \ga to be a literary unity. Yet, such discussions still tend focus on situating \ga within a particular generic classification.%
    \footnote{Notably \cite{bernstein_as2010} and \cite{falk2007}. Cf. \cite{eshel_roitman-etal2011}.}

To illustrate this difficulty, I would like to focus on Moshe Bernstein's treatment of the ``Genre(s)'' of \ga.%
    \footnote{%
        As argued in \cite{bernstein_berthelot-etal2010}. See also the related work of \cite{bernasconi_as2011}.}
    
Bernstein's basic thesis argues that \ga, as a composite work, must be treated as multigeneric, rather than simply as ``rewritten Bible'' or ``parabiblical'' because \ga does not relate uniformly to the biblical text. The difficulty, for Bernstein, comes when one must decide how to characterize the work as a whole. While works such as \jub and Pseudo-Philo could be viewed as works that have been uniformly ``rewritten'' (that is, that the entirety of the work is a single rewriting), works such as \ga (he also includes the \templescroll) could be viewed as ``a series of mini-rewritings of limited scope.''%
    \footnote{\Cite[336]{bernstein_berthelot-etal2010}. I am reminded here of Nickelsburg's similar sentiment regarding the ways that \firstenoch rewrites the flood story several times, arguing that the phenomenon of rewriting moved from smaller units of rewriting to larger, more systematic rewritings. See \cite[89--90]{nickelsburg_stone1984}.}
According to such a characterization, Bernstein writes, ``we have no choice but to refer to Part I [the Lamech and Noah Memoirs] as `parabiblical' and Part II [the Abram Memoir] as `rewritten Bible''' based on the fact that, while the Abram Memoir rewrites portions of Genesis, the Lamech and Noah Memoirs really only take Genesis as a point of departure for their stories and may, in fact, be rewriting other texts.%
    \autocite[337]{bernstein_berthelot-etal2010}
Thus, Bernstein's suggests that \ga does not participate in multiple genres \emph{simultaneously}, but participates in multiple genres \emph{in sequence}.%
    \footnote{This characterization can be contrasted with treatment of the \templescroll as participating in both \rwb and legal genres by Fraade in \cite*{fraade_goldstein-etal2017}.}
To refer to the entirety of \ga as \rwb or as two different kinds of \rwb is, according to Bernstein, unacceptably imprecise. While I am happy to accept a multigeneric characterization of \ga (and any number of other texts), I think Bernstein has sidestepped a more fundamental question by suggesting that the relationship between \ga and its sources is best addressed as an issue of genre. 

The assumption made by Bernstein is that there was a qualitative difference between the sources utilized by \ga,%
    \footnote{While I am sympathetic to viewing \ga as secondary to \jub and \firstenoch, here, I am simply stating this as Bernstein's position.}
an assumption that forms the basis of his characterization of \ga as ``multigeneric.'' This pluriformity is in tension with his larger assertion affirming the unity of the work. It seems to me that the situation may be better analyzed in reverse, namely that the genre of \ga is consistent and it is the assumed qualitative distinction between its sources that should be interrogated. After all, formally speaking, \ga is composed of three (broadly) first-person accounts told from the perspective of three significant patriarchs. In other words, rather than characterizing \ga as a work that utilized both ``biblical'' and ``non-biblical'' sources, it is just as reasonable to begin with the assumption that \ga's method is consistent and that the use of ``non-biblical'' sources actually points to the possibility that \jub and \firstenoch were sources just as legitimate as Genesis. One possible inference from this observation could be that these other works may have been on equal footing with Genesis and enjoyed some special ``scriptural'' (or otherwise authoritative) position for the author of \ga or that such categories were not operative at this time.%
    \footnote{See esp. the work of Eva Mroczek in \cite*[114--155]{mroczek2016}.} To be clear, the terminology of ``\rwb'' is not what is at stake here, but rather the way that we imagine the relationship(s) between \ga and the traditions that inform it.

Although the scholarly consensus since the initial publication of \ga has been that \firstenoch, \jub, and \ga all participate in overlapping or adjacent traditions,%
    \footnote{\Cite[38]{avigad-yadin1956}; \cite[20--22]{fitzmyer2004}; \cite[110--116]{crawford2008}; \cite[8--19]{machiela2009}.}
what remains unclear is the nature and directionality (if any) of these relationships. While Avigad and Yadin suspected that \ga was a source for \firstenoch and \jub,%
    \autocite[38]{avigad-yadin1956}
it is now widely acknowledged that no definitive evidence has yet been assembled to argue one way or another.%
    \footnote{At the risk of over-simplifying the issue, Fitzmyer, Kugel, \vanderkam, and Nickelsburg tend to see \ga as secondary to \jub, while Machiela, Werman, and Segal have argued the reverse. See \cite{vanderkam_feldman-etal2017}, \cite[20--22]{fitzmyer2004}, \cite[174]{nickelsburg2005}, \cite[305--342]{kugel2012}. Cf. \cite{segal_as2010}, \cite[140--142]{machiela2009}, \cite{machiela_mason-etal2012}, and \cite[171--177]{werman_chazon-etal1999}.}

Thinking about \ga in terms of cultural memory means thinking about its composition not simply in source-critical terms, but rather as the synthesis of traditions which, regardless of whether they were considered religiously ``authoritative,'' were operative within the \emph{cultural discourse} of late \secondtemple Judaism. In other words, viewing \ga as the product of cultural memory means taking seriously the idea that the combination of traditions in \ga should not primarily be understood as the genius of an author/editor, but rather that the author/editor should be viewed as discursively engaging with sites of memory and as the instrument by which cultural memory was codified as text. Of course, we must allow for singular, creative contributions of the author/editor of \ga, but even those original contributions should not be treated as if they arose out of a vacuum.%  
    \footnote{Such conscious contributions are examples of memory construction. If from this perspective, one supposed that the author of \ga was responsible for the synthesis of these traditions, \ga would represent another iteration of the process of memory construction and reintegration into the cultural memory of \secondtemple Judaism.}
The cultural memory that surrounded the book of Genesis---the biblical memory related to Genesis---was more broad than the text of Genesis alone and included traditions that we know from \jub and \firstenoch (whether or not they were directly informed by the \emph{texts} of \jub and \firstenoch). The fact that these traditions coexist within \ga speaks to the notion that these traditions engaged with the same sites of memory, which I have called ``biblical memory,'' and that the author of \ga viewed all such sources as useful for his narrative purposes.

\section{Abram in the Diaspora: The Literary Frameworks of \GA}

Having discussed how \ga functioned as the \emph{recipient} of a cultural memory that was broader than the text of Genesis, we may now turn our attention to the ways that \ga was adapted to address its audience within the frameworks of their \emph{social} memory. This section will focus on the way that \ga was shaped by the social frameworks of its compositional milieu, specifically the ways that contemporary cultural discourses and literary forms molded the biblical memory relating to Genesis (specifically, the Abram narrative) into a meaningful participant in the collective memory of \secondtemple Judaism.

As I have already noted, the narrative of \ga is not simply a straight-forward retelling of Genesis from the perspectives of Lamech, Noah, and Abram, but participates more broadly in the biblical memory relating to Genesis (which includes other traditions like those represented in \firstenoch and \jub). However, what is most compelling about \rwb texts very often is the \emph{ways} that they adapt biblical memory. These adaptations can come at the level of story---by adding, removing, or rearranging events---or at the level of narrative discourse by describing events differently or with different emphases.%
    \footnote{Here ``story'' refers to the abstract sequence of actions which the narrative describes. The \emph{way} a story is recounted, on the other hand, is referred to by narratologists as ``narrative discourse.'' Thus, \ga's change from third-person omniscient to a \psgraphical first-person narrative can be understood as a change in narrative discourse which essentially operates on the same story as that of Genesis. See \cite[13--27, esp. 18--19]{abbott2008}.}
In the case of \ga, and in particular in the account of Abram's descent into Egypt in \cols{19}{20}, the biblical narrative has been recast as a first-person Hellenistic novella in a similar vein to other well-known \secondtemple Jewish works such as the narrative portions of Daniel (including the Greek additions), Esther, Tobit, and (arguably) the so-called Joseph novella of Genesis 37 and 39--50.%
    \footnote{See especially Lawrence Wills's work on the Jewish novels and novellas in antiquity: \cite*{wills_jsj2011}, \cite*{wills2002} as well as his important earlier works \cite*{wills1995} and \cite*{wills1990}.}

For the convenience of the reader, I provide here the text and my translation of \ga's account of Abram's descent into Egypt (\q{1}{apGen ar}{} cols. \rom{19}, 10--\rom{20}, 32):%
    \footnote{The Aramaic text is that of M. Abegg in \cite[1:534--537]{perry-tov2014}, in consultation with \cite{fitzmyer2004}. The translation is my own.}

\input{_translations/genapoc_19}
\input{_translations/genapoc_20}

% % TODO: Add  Analysis of Cols. 19-20?
\noindent
Blake Jurgens has recently offered a compelling reading of \ga \cols{19}{20} as a Hellenistic Jewish novella. He argues that the utilization of Hellenistic literary motifs and structures in \ga altered the overall presentation of the pericope for the purpose of edifying Jews living in the Hellenistic world in the shadow of empire.%
    \autocite{jurgens_jsj2018}
Although much of Jurgens's paper is based on long-established observations about the literary influences on \ga, he makes an important discursive turn toward the audience by claiming that \ga was meant to be useful to readers:

\begin{quote}
    By imbuing its story with literary tropes and techniques similar to those found in Dan 1--6, Esther, and other Jewish texts arising out of the Hellenistic period, the author successfully attends to the narratival ambiguities of Gen 12:10--20 through interpretive expansion upon the latent exegetical links of the text while concurrently modifying the narrative to appeal to contemporary literary expectations.\autocite[27]{jurgens_jsj2018}
\end{quote}
\noindent
Thinking in terms of social memory, however, we can appreciate the way that \ga retells the story of Abram is ``remembered into'' the social context of Hellenistic Judaism and is fitted into contemporary social frameworks by the utilization of specific literary forms. The changes that Jurgens identifies as authorial decisions intended to engage with readers can be reframed as \emph{determined by} the social location of the author and the literary tools available to him. In other words, new kinds of literary forms overtake older ones and older stories take on new meanings for new contexts as societies change. This adaptation into new forms is socially determined and should be understood as an example of how social frameworks (in this case, literary frameworks), shape memory  and affect meaning \emph{in the present}.

\subsection{Abram in the Diaspora}

One of the primary features of Jewish Hellenistic novellas is their setting. First, Jurgens notes that these Jewish novellas are typically set in the diaspora, which invariably place the Jewish (or Israelite, in Tobit and Judith's case) protagonist under the hegemony of a foreign power. In the case of \ga, Abram is a sojourner in a foreign land and is under foreign hegemony,  albeit not formally in the ``diaspora.'' Second, from a modern perspective, these stories have a tendency to commit rather egregious factual errors about certain historical particulars such as the names of rulers (Judith 1:1; Dan 4; Tobit) and geographic items (Tobit 5:6). \GA exhibits this feature by utilizing details which almost certainly were inventions of the author or an earlier tradant, such as referring to ``Pharaoh Zoan'' (we know of no such figure) and Hyrcanus (a name popular in the Ptolemaic period, but not attested otherwise), as well as referring to the ``Karmon River'' (probably the Kharma canal), as the one of the seven heads of the Nile river, which it is not.%
    \footnote{\Cite[7]{jurgens_jsj2018}. See also 
        \cite[50--59]{machiela_as2010};
        \cite[197--199]{fitzmyer2004}.}
These details, according to Jurgens, are meant to create a sense of verisimilitude and authenticity within the narrative. Thus, although the story of Abram's sojourn in Egypt as narrated in the biblical text engages with discourses of the \emph{foundation} of Israel, the narrative of \ga seems to be turning the story to engage with the contemporary discourses around the idea of \emph{diaspora}. 

The connection between the contemporary issue of diaspora with the inherited story of Abram's sojourn is another case where keying and framing come into play. In this case, however, it is the story of Abram's sojourn which is, in some sense, ``keyed'' to the issue of diaspora. In other words the way that Abram's sojourn in Egypt was remembered in the \secondtemple period, at least in part, took on new meaning for those sojourning in the diaspora and for those living in the land under foreign hegemony and this reinterpretation of Abram's activities shows through in the literary form that the author of \ga utilized.

\subsection{Abram in the Court of a Foreign King: Literary Genre as Social Framework}

If we place the pericope of Abram's journey into Egypt in \ga under the rubric of diaspora literature, the final scene in the pericope bears a striking resemblance to the so called court contest narratives well known from (especially) the book of Daniel.%
    \footnote{Other court contest narratives include the Joseph Cycle (Gen 41).}
Such narratives, as observed by Collins and others, follow particular narrative progressions with common features.%
    \footnote{\Cite[38--52]{collins1993}; \cite{humphreys_jbl1973}; \cite{collins_jbl1975}; \cite{wills1990}. See also \cite{niditch-doran_jbl1977}.}
Jurgens has convincingly argued that \ga's retelling of Abram's sojourn in Egypt fits such a progression by comparing this pericope to Dan 2, 4, and 5 as well as Gen 41. The resemblance is quite striking. Although based on the earlier work of Collins and Humprheys, Jurgens offers his own outline, which can be summarized as follows:%
    \autocite[21]{jurgens_jsj2018}
\begin{quote}
    1. The foreign king has a problem that he is unable to solve.\\
    2. The king's own personnel are charged with solving the problem\\
    3. The king's personnel are unable to solve the problem\\
    4. The Jewish protagonist is asked to solve the problem\\
    5. The Jewish protagonist is able to solve the problem\\
    6. The Jewish protagonist is rewarded by the king
\end{quote}
\noindent
The biblical account, however, offers a rather anemic description of the events, leaving open the specifics of how Pharoah came to know about Abram and how the monarch was relieved from the plagues. It is easy to imagine how the author of \ga would conceive of Abram's interaction with Pharoah in Gen 12 as analogous to other well-known court contests from Israel's biblical memory, despite the fact that the biblical version offers almost no detail. Gen 12:17--20 reads: 

\subsubsection{Genesis 12:17--20}
\begin{hebrewtext}
    \versenum{12:17}
    ‏וַיְנַגַּע יְהוָה אֶת־פַּרְעֹה נְגָעִים גְּדֹלִים וְאֶת־בֵּיתוֹ עַל־דְּבַר שָׂרַי אֵשֶׁת אַבְרָם׃
    \versenum{18}
    וַיִּקְרָא פַרְעֹה לְאַבְרָם וַיֹּאמֶר מַה־זֹּאת עָשִׂיתָ לִּי לָמָּה לֹא־הִגַּדְתָּ לִּי כִּי אִשְׁתְּךָ הִוא׃
    \versenum{19}
    לָמָה אָמַרְתָּ אֲחֹתִי הִוא וָאֶקַּח אֹתָהּ לִי לְאִשָּׁה וְעַתָּה הִנֵּה אִשְׁתְּךָ קַח וָלֵךְ׃ 
    \versenum{20}
    וַיְצַו עָלָיו פַּרְעֹה אֲנָשִׁים וַיְשַׁלְּחוּ אֹתוֹ וְאֶת־אִשְׁתּוֹ וְאֶת־כָּל־אֲשֶׁר־לוֹ׃
\end{hebrewtext}

\begin{translation}
    \versenum{Gen 12:17}
    Now, \yahweh afflicted Pharoah and his house with terrible plagues on account of Sarai, the wife of Abram. 
    \versenum{18}
    And Pharoah called for Abram and he said, ``What is this that you have done to me? Why did you not tell me that she was your wife? 
    \versenum{19}
    Why did you say, `she is my sister' such that I took her as a wife? Anyhow, here is your wife. Take her and go.''
    \versenum{20}
    And Pharoah gave his men orders about him and they sent him away along with his wife and everything he owned.
\end{translation}

From an innerbiblical perspective, \ga's description of Abram and Pharaoh's interaction might be thought of as a synthesis or exegetical harmonization with the Abimelech doublet in Gen 20, which offers a much more detailed account of Abimelech's confrontation with Abram/Abraham (Genesis 20:8--18):

\begin{hebrewtext}
    \versenum{Gen 20:8}
    וַיַּשְׁכֵּם אֲבִימֶלֶךְ בַּבֹּקֶר וַיִּקְרָא לְכָל־עֲבָדָיו וַיְדַבֵּר אֶת־כָּל־הַדְּבָרִים הָאֵלֶּה בְּאָזְנֵיהֶם וַיִּירְאוּ הָאֲנָשִׁים מְאֹד׃
    \versenum{9}
    וַיִּקְרָא אֲבִימֶלֶךְ לְאַבְרָהָם וַיֹּאמֶר לוֹ מֶה־עָשִׂיתָ לָּנוּ וּמֶה־חָטָאתִי לָךְ כִּי־הֵבֵאתָ עָלַי וְעַל־מַמְלַכְתִּי חֲטָאָה גְדֹלָה מַעֲשִׂים אֲשֶׁר לֹא־יֵעָשׂוּ עָשִׂיתָ עִמָּדִי׃
    \versenum{10}
    וַיֹּאמֶר אֲבִימֶלֶךְ אֶל־אַבְרָהָם מָה רָאִיתָ כִּי עָשִׂיתָ אֶת־הַדָּבָר הַזֶּה׃
    \versenum{11}
    וַיֹּאמֶר אַבְרָהָם כִּי אָמַרְתִּי רַק אֵין־יִרְאַת אֱלֹהִים בַּמָּקוֹם הַזֶּה וַהֲרָגוּנִי עַל־דְּבַר אִשְׁתִּי׃
    \versenum{12}
    וְגַם־אָמְנָה אֲחֹתִי בַת־אָבִי הִוא אַךְ לֹא בַת־אִמִּי וַתְּהִי־לִי לְאִשָּׁה׃
    \versenum{13}
    וַיְהִי כַּאֲשֶׁר הִתְעוּ אֹתִי אֱלֹהִים מִבֵּית אָבִי וָאֹמַר לָהּ זֶה חַסְדֵּךְ אֲשֶׁר תַּעֲשִׂי עִמָּדִי אֶל כָּל־הַמָּקוֹם אֲשֶׁר נָבוֹא שָׁמָּה אִמְרִי־לִי אָחִי הוּא׃
\end{hebrewtext}

\begin{translation}
    \versenum{Gen 20:8}
    Abimelech rose early in the morning and called all his servants and spoke all these things in their ears. And the men were very frightened. 
    \versenum{9}
    So Abimelech summoned Abraham, and said to him, ``What have you done to us? How have I sinned against you such that you have brought terrible guilt upon me and my kingdom? You have done things to me which should not be done.''
    \versenum{10}
    And Abimelech said to Abraham, “What were you thinking that made you do this thing?” 
    \versenum{11}
    Abraham said, ``I thought, `Surely there is no fear of God at all in this place, and they will kill me because of my wife. 
    \versenum{12} 
    Also, honestly, she is my sister; the daughter of my father but not the daughter of my mother and she \emph{became} my wife.
    \versenum{13}
    And when God caused me to wander from my father's house, I said to her, `This is the kindness you must do for me: at every place when we enter there, say  ``He is my brother."'"
\end{translation}

\begin{hebrewtext}
    \versenum{Gen 20:14}
    וַיִּקַּח אֲבִימֶלֶךְ צֹאן וּבָקָר וַעֲבָדִים וּשְׁפָחֹת וַיִּתֵּן לְאַבְרָהָם וַיָּשֶׁב לוֹ אֵת שָׂרָה אִשְׁתּוֹ׃
    \versenum{15}
    וַיֹּאמֶר אֲבִימֶלֶךְ הִנֵּה אַרְצִי לְפָנֶיךָ בַּטּוֹב בְּעֵינֶיךָ שֵׁב׃
    \versenum{16}
    וּלְשָׂרָה אָמַר הִנֵּה נָתַתִּי אֶלֶף כֶּסֶף לְאָחִיךְ הִנֵּה הוּא־לָךְ כְּסוּת עֵינַיִם לְכֹל אֲשֶׁר אִתָּךְ וְאֵת כֹּל וְנֹכָחַת׃
    \versenum{17}
    וַיִּתְפַּלֵּל אַבְרָהָם אֶל־הָאֱלֹהִים וַיִּרְפָּא אֱלֹהִים אֶת־אֲבִימֶלֶךְ וְאֶת־אִשְׁתּוֹ וְאַמְהֹתָיו וַיֵּלֵדוּ׃
    \versenum{18}
    כִּי־עָצֹר עָצַר יְהוָה בְּעַד כָּל־רֶחֶם לְבֵית אֲבִימֶלֶךְ עַל־דְּבַר שָׂרָה אֵשֶׁת אַבְרָהָם׃  
\end{hebrewtext}

\begin{translation}
    \versenum{Gen 20:14}
    Then Abimelech took sheep and cattle, and male and female slaves, and gave them to Abraham, and restored his wife Sarah to him.
    \versenum{15}
    Abimelech said, ``See, my land is before you. Settle where it seems good to you.'' 
    \versenum{16}
    But to Sarah he said, ``Look, I have given your brother a thousand pieces of silver; it is your exoneration before all who are with you; you are completely vindicated.'' 
    \versenum{17}
    Then Abraham prayed to God; and God healed Abimelech, and also healed his wife and female slaves so that they bore children. 
    \versenum{18}
    For \yahweh had completely closed all the wombs of the house of Abimelech on account of Sarah, Abraham's wife.
\end{translation}
\noindent
While the Gen 12 account is very terse, the Gen 20 account includes a dream-revelation (20:6--7), specifies that the plagues that afflicted the monarch impeded his sexual activities (specifically with Sarah), and describes Abraham praying over Abimelech and his household to heal them. Similar details are given in \ga account which likewise includes a dream-revelation, notes that the plagues were sexual in nature, and describes Abram praying over Pharoah and his household for healing. 

While these similarities may indeed represent some kind of literary conflation between the two accounts,%
    \footnote{From a memory perspective I would prefer to account for \ga's adoption of certain details from Gen 20 in more passive terms where the specifics of the Gen 12 story are, where absent, supplied from another well-known, typologically similar, source. It is also, perhaps, worth noting that this doublet has traditionally been attributed source critically to different hands (Gen 12 = J; Gen 20 = E). See \cite[15]{driver1956}.} 
at the level of genre and structure, conflation with Gen 20 cannot account for \ga's reframing as a court-contest. For example, the dream-revelation in Gen 20 is given to Abimelech, rather than to Abram as in \ga. Moreover although Abraham prays for healing for Abimelech and his household in a very similar fashion to the way he is portrayed in \ga praying for Pharoah, in Gen 20, he does so only after Abimelech effectively ``pays him off.'' It is the revelation given to Abimelech in a dream which causes him to ``repent'' in Gen 20, while in \ga, the miraculous healing of Pharoah and his household functions as the sign and catalyst for Pharoah's rich rewarding of Abram. Although this difference may seem subtle, the primary feature of the court-contest is the demonstration of God's power through the protagonist which leads to the foreign king's repentance/conversion and the rationale for his rewarding of the protagonist. In other words, while it may have been that \ga used details from Gen 20 to supplement the account from Gen 12, \ga's framing of Abram's contest with Pharoah cannot be solely attributed to a harmonization of the Gen 12/20 doublet. Thus, drawing on details from, or perhaps just inspired by, the Abimelech doublet in Gen 20, the author of \ga was able to reframe this portion of the Abram narrative to conform to the common court-contest pattern, which, as Jurgens rightly notes, surely would have been an effective and entertaining adaptation by comparison to the account from Genesis.

\subsection{Other Literary features and Motifs: Abram as Oracle}
Although the Abimelech story in Gen 20 includes a dream-revelation, it is noteworthy that in \ga, Abram himself is given the dream as a means of warning him about how the Egyptians would attempt to kill him on account of Sarai's beauty. Where the biblical text credits Abram's intuition for anticipating the Egyptians' desire for Sarai (though we are left to wonder whether he would have been killed had the ruse not been realized), \ga describes Abram receiving a portentous dream vision characteristic of other \secondtemple literature.%
    \footnote{\Cite{gevirtz_maarav1992}; \cite[184]{fitzmyer2004}; \cite{dacy_tzoref2013}.}

Although dream-visions are not unique to the \secondtemple period, their ubiquity within Jewish literature from the \secondtemple period is indisputable. In his treatment of the Dream-Visions among the Aramaic \dss, Andrew Perrin describes both Abram and Noah as being ``recast as a dreamer[s]'' within \ga.%
    \footnote{\Cite[52--57]{perrin2015}. See also 
        \cite{eshel_klostergaard-etal2009} and 
        \cite{machiela_falk-etal2010}.}
While Noah is not described as a dreamer within the biblical text, within \ga, he seems to have been the recipient of as many as five dream-visions.%
    \footnote{\Cite[53]{perrin2015}. Elsewhere in the enochic literature dreams weight heavily in the events surrounding the flood, if not always given to Noah (\firstenoch, Book of Giants, etc.).}
Restricting the discussion to Abram, however, Perrin suggests that the insertion of a dream-vision into the story on the eve of Abram and Sarai's descent into Egypt functioned as part of a larger project to ``extend Abram's prophetic credentials in light of Gen 20:7.''%
    \autocite[55]{perrin2015} 

Fitzmyer notes that the component parts of this dream---``cedar'' (Aramaic \aram{ארז}) and ``date-palm'' (Aramaic \aram{תמרא})---are derived from Ps 92, which declares ``the righteous will flourish like the date palm (Hebrew \he{כַּתָּמָר}); like a cedar (Hebrew \he{כְּאֶרֶז}) in Lebanon he will grow'' (Ps 92:13). The identification of Abram and Sarai with the cedar and date-palm, respectively, is plain enough by the parallel to what happens next in the narrative. The interpretation is supported by the grammatical gender of the terms \aram{ארז} (masc.) and \aram{תמרא} (fem.) which correspond to the gender of the characters. Although grammatical gender does not correspond to natural gender, the identification of Sarai with the date-palm is supported further by the use of ``Tamar'' as a feminine personal name within the Bible (Gen 38:6; 2 Sam 13:1; 14:27).

The dream itself provides an allegorical vision that credits the date-palm (Sarai) with saving the cedar (Abram) from the people seeking to destroy it. Although there is a question whether the beginning of \col{19}, 14--15 should read \aram{והא אזר חד ותמרא כחדא צמחו מן שרש חד} ``a cedar and a date-palm \emph{growing from a single root}'' (so DJD), or \aram{ והא אזר חד ותמרא חדא יאיא שגי} ``a cedar and a date-palm [which was] \emph{very beautiful}'' (so Fitzmyer), all editions understand \col{19}, 16 to read \aram{ארי תרינא מן שרש חד צמחנא} ``for the two of us grow from a single root.'' Thus, the purpose of the dream  is to show Abram the way that he should avoid being ``cut down and uprooted'' by the Egyptians, namely, by claiming that he and Sarai had ``sprung from the same root,'' viz. are siblings. This interpretation is also offered by Abram himself in \col{19}, 19--21.

Significantly, the later \GenRabbah connects the this section of Genesis with Ps 92 and utilizes the cedar/date-palm imagery as well, albeit in a different manner. Specifically, during its treatment of the description of the plagues which God inflicted on Pharoah, \GenRabbah begins with a citation of Ps 92, ``The righteous will flourish like the date-palm (Hebrew \he{תמר}); like a cedar (Hebrew \he{ארז}) in Lebanon he will grow'' and this comparison to date-palms and cedars makes several digressions. First, building on the idea of righteousness, \GenRabbah observes that both cedars and date-palms are ``straight'' trees, largely without crooks and crotches.%
    \footnote{Although the Psalm uses the typical term for ``righteousness'' (Hebrew \he{צַדִיק}), another common biblical term for a person who acts in an upright manner is ``straight'' (Hebrew \he{יָשָׁר}). The author of \GenRabbah seems to be playing on this association. Furthermore, according to \GenRabbah, tall trees cast long shadows; the length of these shadows represents the fact that the reward for such righteous people will only come later.}
The second digression focuses on the ability of date-palms to produce fruit (including through grafting) and the usefulness of, especially, date-palms for all manner of practical concerns. \GenRabbah then extends the comparison to the whole of Israel:
\begin{aramaictext}
    מַה תְּמָרָה זוֹ אֵין בָּהּ פְּסֹלֶת אֶלָּא תְּמָרֶיהָ לַאֲכִילָה וְלוּלָבֶיהָ לְהַלֵּל חֲרָיוֹת לְסִכּוּךְ סִיבִים לַחֲבָלִים סַנְסַנִּים לִכְבָרָה שִׁפְעַת קוֹרוֹת לְהַקְרוֹת בָּהֶם אֶת הַבַּיִת כָּךְ הֵם יִשְׂרָאֵל אֵין בָּהֶם פְּסֹלֶת אֶלָּא מֵהֶם בַּעֲלֵי מִקְרָא מֵהֶם בַּעֲלֵי מִשְׁנָה מֵהֶם בַּעֲלֵי תַּלְמוּד מֵהֶם בַּעֲלֵי הַגָּדָה
\end{aramaictext}
\begin{translation}
    As no part of the palm has any waste, the dates for eating, the branches for Hallel, twigs for woven coverings, fibrous bits for ropes, leaves for baskets, planed boards for ceiling rooms. Likewise concerning Israel, there are no worthless people among them: some masters of Scripture, some masters of Mishnah, some masters of Talmud, some masters of Haggadah. (\genrab{41.1})
\end{translation}
\noindent
The final comparison makes the claim that, like dealing with Israel, climbing these tall trees is perilous. The proof, for \GenRabbah, brings us back to the verse at hand. That Pharaoh was plagued by Yahweh when he took Sarai for himself demonstrates the danger in engaging with Israel as an adversary. What is significant here is that the authors of both \ga and \GenRabbah connect Ps 92 with this section of Genesis, but importantly, they do so with different sets of interpretive ``evidence.'' In other words, the two texts agree on the \emph{fact} of the connection between Ps 92 and the story in Gen 12, but they arrive at their interpretations through different means.%
    \footnote{Perhaps based on the Psalm's later reference to bearing children in one's old age:
    \begin{quote}
        Planted in the house of Yahweh; they will flourish in the courts of our God\\
        They will still bring forth fruit in old age; they will be full of sap and green\\
        (Ps 92:14--15)
    \end{quote}.}
Yet, the fact that Ps 92 and Gen 12 are explicitly connected in both \ga and \GenRabbah seems more than a coincidence. Thus, while \ga's use of the cedar/date-palm imagery may rely on some previous tradition, the dream revelation itself is best understood as an example of the author of \ga utilizing the literary tropes of his own time and place. 

% \subsubsection{Abram as Sage}
% [TODO: A Brief discussion of Abram being depicted as a sage (sought for knowledge; reads from the ``book of the words of Enoch,'' etc)]

% \subsubsection{A \emph{waṣf} about Sarai}
% [TODO: Talk about Waṣf]

% \subsection{Conclusions}

The recasting of Abram's sojourn as ``diaspora,'' his conflict with Pharaoh as a court-contest along with the portrayal of Abram and Noah as dreamers can be understood in terms of social memory, as the author of \ga pressing the stories of Genesis into existing literary genres. Insofar as ``genres'' can be understood as commonly understood literary conventions---a social ``contract'' of expectations between the author and her audience---they are socially defined and, for our purposes, function as what \halbwachs would call ``social frameworks.'' As Abram's sojourn in Egypt could take on new meanings within the context and social framework of diaspora Judaism during the \secondtemple period, so too the common trope of the court-contest---well-known from the book of Daniel---provided a new framework into which the story of Gen 12 could be read. Thus, Jurgens's basic premise---that these stories are ``updated'' for a new audience---takes for granted what the memory approach makes explicit: \secondtemple Jews had their own ways of thinking about the way that God interacted with the ancients, and how pious Jews acted in particular circumstances. These social frameworks provided new structures for understanding the stories that they inherited from the biblical tradition. Thus, rather than only thinking about how the author was trying to ``fix'' the biblical account, from the memory perspective we can imagine the author of \ga not only interpreting the biblical tradition, but making efforts to contextualize it within his own literary frame of reference.

\section{\ga as Pseudepigraphon}

% Intro about third level of discourse
While \ga can be seen engaging with its received cultural memory through its sources and engaging with its contemporary social memory at the level of literary form and genre, \ga also participates in the construction of cultural memory for transmission to future generations. Although \emph{all} literary and cultural products can participate in constructing cultural memory, in this section, I will argue that \ga's \psgraphic form participates in this constructive act differently than other forms of literature, in particular the biblical text.% TODO: What does this mean? Tease all of this out.
    \footnote{I reiterate that although the term ``biblical'' is anachronistic for the late \secondtemple period, it is a usefully concise term for my purposes.}

\subsection{The Hebrew Bible as a Baseline}

The vast majority of the Hebrew Bible is narrated in the third-person omniscient and is formally anonymous. Although there are exceptions to this generalization (most notably within the prophetic corpus, such as Isa 6--8, the so-called Nehemiah Memoir (Neh 11--13), and perhaps works such as Deuteronomy and Song of Songs), but for the lion's share of the biblical text, the author (and narrator) operates invisibly.

The rhetorical force of this particular authorial voice, as observed by Erhard Blum, is significant for the function of the Hebrew Bible's participation in the collective memory of the communities that claim it as their own. Although the implied author does occasionally engage directly with the reader by offering explanatory observations (for example where the author inserts phrases like ``this is why\ldots{}'' or ``\ldots{}until this day''), for all intents and purposes, the author presents as both \emph{reliable} and \emph{authoritative} without a hint of subjectivity. As Blum puts it, ``In this sense the narrative does not distinguish the depiction from the depicted.''%
    \autocite[33]{blum_barton-etal2007}
Put another way, the text does not acknowledge that it \emph{has} an author, it simply \emph{is}. The rhetorical effect of this invisible, omniscient author is to collapse the knowledge gap between the reader and the events narrated by removing the author from view. This move, according to Blum, allows the text to convey ``an unmediated truth claim which is not based on the author's distinguishable critical judgments and convictions.''%
    \autocite[33]{blum_barton-etal2007}
The effectiveness of this implied author, according to Blum, is tied to the pragmatics of the text, that is, tied to the context of the biblical narratives as scripture (though Blum does not refer to ``scripture'' \emph{per se}). The implied audience of the biblical narratives by-and-large can be understood as group-insiders for whom the biblical text worked to reinforce group identity.

Of course, the ``unmediated truth claims'' of the biblical text \emph{were}, in fact, mediated and reinforced by those who (orally or otherwise) transmitted the tradition from one generation to another.%
    \autocite[33]{blum_barton-etal2007}
Individuals within the community---teachers and religious leaders and even parents---become the voice of the biblical text as it is passed on. In other words, one might say that the narrator of the biblical text is the community itself---its collective memory. Blum writes:

\begin{quote}
    If we assume that the traditional literature was primarily transmitted through oral means, then the narrator who is speaking supplies the material with a personal presence; he is not present as an author who judges and evaluates his sources from a critical distance, but as a `transmitter' who participates in the tradition itself and is able to lend it credence through his own personality, his standing, and/or his office.\autocite[33]{blum_barton-etal2007}
\end{quote}
\noindent
In other words, the authoritative claims of ``biblical'' texts are actually made by their communities and not by the text itself. Thus, the way biblical texts participate in the collective memory is determined by their \emph{use}---how their \emph{readers} frame their function and how the text relates to the collective memory. 

\subsection{On \Psy and the \Psa}

Because significant portions of \ga are written in the first person as though written by Lamech, Noah, and Abram, \ga may be formally included in the literary category of \psy.%
    \footnote{%
        On the broader topic of first-person discourse among the \dss, see 
        \cite{stuckenbruck_roitman-etal2011}. See also his work on the Epistle of Enoch \cite*{stuckenbruck_dsd2010}.}
Before we move on, however, it is worth taking a moment to clearly define what is meant by ``\psy,'' ``\psa,'' and related terms.%
    \footnote{The topic of \psy has received a large amount of very sophisticated attention in recent years. See especially
        \cite{mroczek2016}
        \cite{tigchelaar_tigchelaar2014}
        \cite{reed_towsend-moulie2011}
        \cite{reed_jts2009}
        \cite{reed_ditomasso-turcescu2008}
        \cite{najman_hilhorst-puech2007}
        \cite{najman2003}.
        In the simplest terms, \psa are texts which are fictively purported to be written by figures (typically) from the ancient past.}

The ancient use of the term \psa denoted spurious texts which Church leaders believed to be intentionally misleading about their authorship.%   
    \footnote{See esp. Eusebius's \emph{Hist. eccl.}~6.12.2 where the Bishop of Antioch, Serapion, refers to the \emph{Gospel of Peter} among the a number of works ``falsely attributed'': \gk{γάρ, ἀδελφοί, καὶ Πέτρον καὶ τοὺς ἄλλους ἀποστόλους ἀποδεχόμεθα ὡς Χριστόν, τὰ δὲ ὀνόματι αὐτῶν   ψευδεπίγραφα ὡς ἔμπειροι παραιτούμεθα, γινώσκοντες ὅτι τὰ τοιαῦτα οὐ   παρελάβομεν}. ``For we, brothers, accept both Peter and the other apostles as Christ, but we skillfully reject those falsely ascribed writings, knowing that they were not handed down to us.''}
Thus, the term has tended to carry a somewhat negative connotation, even when such a connotation is not warranted. Implicit in the negative use of the term is the assumption that ``false'' attribution was malicious, or at the very least intentionally misleading. Yet, the number of (esp.~Jewish) \psgraphical texts discovered within the past century provides good reason to question the assumption that pseudonymous authors' intentions were to deceive their readers. On the contrary, the sheer number of \psgraphical works now known to us suggests that the historical reality and social function of \psgraphical works was not simply a matter of being ``falsely attributed.''%
    \footnote{\Cite[53--58]{mroczek2016}. See also \cite{reed_jts2009}.}

At the other end of the spectrum, because so many early Jewish texts seem to fall into the category of \psa, in some scholarly discourse, the term ``\psa'' has become generalized to encompass any text written in around the turn of the era which did not make it into the canon of rabbinic Judaism or early Christianity. Bernstein observes, for example, that although the first volume of James Charlesworth's two-volume \emph{Old Testament Pseudepigrapha} contains a number of formally \psgraphic works, the second volume includes many which do not meet the formal definition of \psa.%
    \footnote{\Cite[2]{bernstein_chazon-etal1999}. See also \fullcite{charlesworth_OTP}.}
This expansive practice is not particularly helpful for clarifying the term and so I will attempt to restrict my usage to a more clearly defined set of criteria.

Moshe Bernstein, in his discussion of the phenomenon of \psy, distinguishes between ``authoritative'' \psy and ``decorative'' \psy.%
    \footnote{He also identifies a third form, ``convenient'' \psy, which is located somewhere between the two. \cite[3--7]{bernstein_chazon-etal1999}.}
By ``authoritative'' \psy, Bernstein refers to texts that \emph{portray themselves} as being written by a particular figure. Portions of \firstenoch (in particular the latter three books, Astronomical Writings [72--82], Dream Visions [83--90], and the Epistle of Enoch [91--107]), which present themselves as if they were written by Enoch himself, are prime examples of ``authoritative'' \psy. \GA, too would fall into this category. Psalm 23, on the other hand, although attributed to David, was presumably not \emph{actually} written by David. Moreover, whoever did write Ps 23 (again, presumably) did not intend to write it \emph{as if} it had been written by David. Rather, the Psalm was simply \emph{attributed} to David, along with many others, in part due to the tradition of David being a musician.%
    \footnote{See, for example 2~Chr~23:18 and Ezra~3:2,10. \cite{mays_interpretation1986}; \cite{sarna_stein-loewe1979}.}
Thus, Ps 23 could be classified as ``decorative'' \psy. Thus, the difference between ``authoritative'' and ``decorative'' \psy can, in some sense, be boiled down to the notoriously difficult issue of authorial intent---whether a text was \emph{intended} to be read as \psa or whether the work was anonymous and only later attributed to an explicit author.

Less clear-cut examples, however, require a more nuanced treatment. For example, Deuteronomy is not generally referred to as among the \psa, yet, from a literary perspective, it is framed as \he{הדברִם אשׁר דבר משׁה אל־כל־ישׂראל} ``the words which Moses spoke to all Israel'' (Deut 1:1a). Although the whole narrative is not written in the first person, long sections of the book are treated as verbatim recountings of Moses's speech. Was Moses the author of Deuteronomy? Most critical scholars have dated Deuteronomy to the late monarchic period and thus have eschewed the traditional mosaic attribution. But whether Deuteronomy was \emph{written} as \psa or just attributed to Moses after the fact is difficult to say with certainly and the matter is further complicated by the editorial processes that the book likely underwent through the centuries.%
    \autocite[143--172]{toorn2007}
What we \emph{can} say is that there are concrete literary cues within Deuteronomy which make the attribution to Moses easier. Framing Deuteronomy as ``the words which Moses spoke,'' while not formally ``\psa,'' participates in the construction of memory in a similar fashion as \psa proper.

\subsection{Pseudepigrapha, \ga, and Memory Construction}

If we take seriously Blum's characterization of the way that the anonymous, third-person omniscient biblical text may have engaged with the collective memory of Israel based on formal, narratological features within the text, it stands to reason that \ga as first-person \psy would engage that collective memory in a different way, despite the fact that some of the stories within \ga are found in the book of Genesis. In other words, the literary form of \ga affects how it relates \emph{back} to the biblical memory, and how it can be used in the further \emph{construction of} that memory.

The \psgraphic quality of \ga shapes the way that the text engages with the remembered past by describing the biblical story through the mouths of important figures. This explicitness changes the way that the reader understands how the text fits into the collective memory by shifting the locus of authenticity onto the text's putative author and away from the mediating figures within the community. In other words, as an example of \psy, \ga can be thought of as a set of fictional \emph{primary sources} that bypass the received tradition. As these sources are used and enter into the discourse of the broader biblical memory, they are able to function not simply as ``alternate'' versions of events but as qualitatively distinct contributions to the tradition as it is passed on to the next generation.%
    \footnote{By analogy to Hindy Najman's notion of ``Mosaic Discourse,'' here I am saying that \ga is participating in a broader ``biblical'' discourse insofar as it participates in discourses surrounding Lamech, Noah, and Abram. See \cite[1--40]{najman2003}.}

Of course, referring to \psy as ``fictional primary sources'' may overstate these texts' importance or otherwise misunderstand how ``authentic'' these texts were thought to be by various and sundry religious groups in antiquity. On the one hand, it could be that readers understood that such novel fictional adaptations took certain artistic license with their biblical \emph{Vorlagen}. By way of analogy, modern adaptations of literary works are expected to deviate to a certain degree from their source material. Readers often expect a degree of ``dramatization'' and are clued into these practices by convention. Such adaptations are not typically understood to be attempts to replace the source material, but are meant to make otherwise obscure or difficult works more accessible to new audiences. On the other hand, there certainly are examples of \psgraphical texts which ultimately \emph{did} become authoritative for certain religious groups.%
    \footnote{For example, the Ethiopian Orthodox Church includes \firstenoch among its scriptures. Tobit, too may, under certain rubrics, be considered \psa, and is included within the Roman Catholic and Eastern Orthodox deuterocanons. Insofar as deutero- and trito- Isaiah were penned as is written by Isaiah, they too could be considered \psa. It is also likely that a number of the so-called disputed Pauline letters within the Christian New Testament were not penned by Paul and are properly \psgraphical.}
My point here is not to suggest that there were multiple ways to understand \psgraphical writing in antiquity so much as to point out that discussions of ``false'' or ``authentic'' attribution are generally from later periods and do not tell us anything meaningful about \emph{why} such a text was written or \emph{how} it would have been understood by its original readers.

\GA was never considered ``scripture'' so far as we know, but that does not mean that it did not participate in the broader biblical memory, even if only in the popular imagination. But even at the level of the popular imagination---even as an entertaining fiction---\ga participated in how its society conceived of the Genesis narratives. Regardless of whether the memoirs in \ga were thought to be ``authentic,'' they represent both an interpretive understanding of biblical memory and an original contribution to that memory.

\section{Conclusion}

As I have demonstrated, \ga may be understood to have taken part in three discrete mnemonic processes: 1) the reception of cultural memory, 2) the reshaping of memory by contemporary social frameworks, and 3) the active construction, codification, and reintegration of memory for future transmission. 

First, \ga functions as the recipient of cultural memory through its engagement with what I refer to as biblical memory. I argued that \ga drew from more than just the biblical text and instead drew from a whole constellation of traditions and stories surrounding the early figures of Lamech, Noah, and Abram. Although the nature of the relationship(s) between \firstenoch, \jub, and \ga is not well understood, what is clear is that the cultural memory that surrounded the book of Genesis---the biblical memory related to Genesis---was more broad at the time that \ga was composed than simply the text of Genesis. The cultural memory from which \ga drew included additional traditions adjacent to the text of Genesis that we know from \jub and \firstenoch whether or not \ga itself drew from the \emph{texts} of \jub and \firstenoch, or vice versa.

% Second, I will discuss the ways that \ga was shaped by the social frameworks which inherited it through a discussion of literary genre and shared formal characteristics with contemporary texts.

But the presentation of these traditions was not a straight-forward synthesis of their content. The author of \ga utilized generic and thematic elements common to the social location in which it was written. Although the account of Abram's encounter with Pharoah in Gen 12 is a rather anemic narrative, \ga does not simply fill out missing details but recasts the final confrontation as a court-contest in the tradition of Daniel and Joseph. Even the Abraham/Abimelech doublet in Gen 20, although a more detailed narrative, cannot account for this transformation. Instead, I have proposed that the utilization of the court-contest as well as the depiction of Abram as a dreamer and his sojourn as diaspora were ways for the author of \ga not only to make his narrative entertaining, but to fit it into the extant social frameworks (read: genres) of the late \secondtemple period.

Finally, I discussed how \ga participated in the construction of cultural memory through its use of \psgraphical discourse. By participating in the genre of \psy, the author of \ga engaged in the further construction of biblical memory by presenting the text of \ga as a first-person narrative. Although we cannot know specifically how \ga was received by its audience in antiquity, the fact that it presents as a ``primary source'' for the stories of Genesis (or, more precisely, the stories which participate in the biblical memory related to Genesis) was meant as a cue to the reader for how to understand \ga's claim to authority, whether that claim was minimal (as with a modern film adaptation where the audience expects certain artistic license) or genuinely intended to co-opt the authority of its pseudonymous author (as with Paul's disputed letters).

As a product of memory, \ga fits this three-fold schema well. Treating \ga simply or even primarily as a way of explaining the book of Genesis does not do justice to the complex and varied processes and traditions that informed the production of \ga or adequately account for the plurality of purposes for which \ga could have been intended. Moreover, reading \ga through the lens of memory allows us to treat the work \emph{as a whole} without the need to explicitly designate the relative status of its putative sources. Instead, memory studies offers a way to talk about how the author of \ga was able receive, recontextualize, and codify the traditions about Genesis (remembered) for himself and his contemporaries and how these processes participated in the construction of Jewish identity during the \secondtemple period.

In the next chapter I will address the ways that \rwb participated in ``imperative discourses'' through a discussion of the book of \jub and will attempt to advance the conversation beyond \emph{abstract} notions of ``the remembered past'' toward treating memory construction as a means to affect concrete \emph{practice}.