
\chapter{Genesis Pseudepocryphon}

 This chapter examines the \ga through \ldots{} and it will do so by discussing the \ga at three distinct levels of discourse: structurally at the macro level as an example of \psy, in terms of keying and framing at the level of narrative discourse, and finally from the perspective of biblical discourse.

 \section{\GA, First Person Narrative, and \Psy}

 One of the most striking features of the \ga when compared to other \rwb texts is its pervasive use of the first-person voice to narrate the events of its story. This quality sets \ga apart from the majority of narrative material in the Hebrew Bible which, with notable exceptions, usually maintains an omniscient third-person voice. The \ga's use of first person must be nuanced, however, by the fact that the work presents itself as a collection of first person ``memoirs''\footnote{I will use the term ``memoir'' throughout this chapter as a way of referring to the distinct (mostly) first-person narratives found in the \ga. This is simply a convenience term that highlights the formal characteristic of being written in the first person voice without any reference to the authenticity of the work and in alignment with the convention of referring to first-person narratives in the Bible as ``memoirs'' (e.g., the ``Nehemiah Memoir'' or the ``Isaiah Memoir'').} from three of the Patriarchs from Genesis (Lamech, Noah, and Abram). So, while each memoir does indeed utilize the first person, the narrator itself changes throughout the course of the text.
 
Speaking of the \ga as a single text, however, should not be taken for granted. Therefore, it is necessary for the moment to consider whether the \ga should be considered ``a'' text or whether instead it should be treated as a collection of ``texts.''  Although the composite character of the \ga was noted in the \emph{editio princeps} by Avigad and Yadin, they maintained that \ga functioned as a single literary unit though allowing that it was made up of several literary sources.\autocite[38]{avigad_yadin1956} Since then, however, the unity of the work has been further interrogated and analyzed from a number of perspectives and the unity of \ga is more tenuous than ever.

Perhaps the most compelling argument for the genuinely composite nature of \ga has been offered by Moshe Bernstein, who has pointed out that discrete units of the \ga utilize distinct titles and epithets for \yahweh.\footnote{\cite{bernstein_jbl2009}. It is delightfully reminiscent of the classical formulations of the Documentary Hypothesis. Attributed to Graf and Wellhausen.  See also~\cite{bernstein_as2010} and~\cite{weigold_as2010}} He notes, for example, that...

% TODO: Give details of Bernstein's argument, 
% TODO: Caveat Bernsteins arguement; note Bernstein's own "other" approach

From a structural standpoint, it is not at all clear whether these three ``memoirs'' are meaningfully related and any thematic consistency is readily explained through the collating process itself.\footnote{In other words, we could easily suppose that the reason that three ``texts'' such as these would be grouped together on a single scroll was that they shared certain common themes or formal characteristics.} One could certainly imagine the \ga as a collection of fictional patriarchal memoirs collected onto a single scroll, and organized roughly by each text's correspondence to the chronology of the biblical narrative. 

In fact, this is precisely how the \ga presents itself. Although the beginnings of the Lamech and Abram accounts are lost due to damage to the scroll, col. 5.29 seems to offer a superscription for the account of Noah, reconstructed as [\emph{pršgn}] \emph{ktb mly nwḥ} ``[a copy of] the book of the words of Noah.'' Based on this superscription, it is reasonable to suppose that the accounts of Lamech and Abram, too, had such superscriptions, although at least two complicating factors should be taken into account. First, while the extant portions of \ga present themselves as distinct units, both the very beginning and the very end of the scroll have been lost. From a structural standpoint, this fact should elicit caution because it is at the beginning and end of texts where such features as ``framing narratives'' and other explanatory material is often located. Without definitive evidence of the presence or absence of such features one must be extra cautious when making observations about the rhetorical purpose of the macro structure of a particular text. Second, although all three accounts are \emph{generally} written in the first-person, as noted above, none of them are rigorously committed to maintaining the voice. As Loren Stuckenbruck notes, each of the three ``documents,'' at one point or another, falls into some kind of third-person voice: Lamech in 5.24--25, Noah in 16.14--17.19, and Abram in 21.23--22.34. Curiously, Stuckenbruck includes the superscription(s) as examples of this inconsistency and does not distinguish between instances where the narrator moves into the third person \emph{within} the narrative and cases where one might suppose the presence of an editorial voice.

The problem of whether to understand the superscription(s) as ``internal'' to the work is a good example of how the macro-structure of the work is important for this kind of analysis. By treating the superscription as a contribution of the ``author'' of a unified \ga, Stuckenbruck understands the superscription to be ``in the third person'' and would (apparently) treat each first-person account as an embedded narrative within a larger framing narrative (of which the superscription would be a part). For example, if the beginning of the scroll gave a brief framing narrative, describing a young man who discovered three scrolls in the desert and thus proceeded to provide ``a copy of the book of the words of X,'' Stuckenbruck would be absolutely correct. However, if one understands the superscription to be an editorial insertion, it does not make sense to include it as an example of third-person discourse for the same reasons it does not make sense to say that Ps 23 uses third-person discourse by beginning with \emph{mizmôr lə-dāwid}.\footnote{\cite[315--316]{stuckenbruck_roitman-etal2011}. See also \cite[15--16]{bernstein_chazon-etal1999}. Even supposing a single author for \ga, as Stuckenbruck and others imply, I am still inclined to consider the superscriptions separately from the former examples because they would exist outside the frame of each embedded narrative.}
In other words, the way that the \ga presents itself, I believe, should best be understood as a \emph{collection} of memoirs compiled by an editor who would have, putatively, supplied a set of paratextual superscriptions.

The fact that \ga presents itself as a collection of disparate texts, however, does not mean that we must treat it as such. Historical and literary disjunction aside, the fact that the \ga was preserved as a ``collection'' tells us something about how its early readers may have understood the relationship between its (putative) constituant parts. Although early readers may not have understood \ga to be a unity, placing each ``part'' of the \ga into a collection queues the reader to engage each part with the same (or similar) reading strategies. 
% TODO: Get something on canonization/collection as it relates to genre?
That means that regardless of the original intent of each of \ga's constituant parts---and therefore whether or not one considers it to be a literary unity---we can still fruitfully consider \ga as a whole with respect to how it may have functioned for its readers and, therefore, as an object of social memory.

 
 \subsection{Pseudepigraphy and the Implied Author}

 Regardless of \ga was composed by a single person or several, the fact that much of the text was written in the first person voice of early biblical figures formally places \ga into the literary category of \psy. Because the terms \psy and (perhaps moreso) \psa are commonly used  in the field of \secondtemple studies to speak generally about ``non-canonical'' and ``fictional'' works I would like to take a moment and make explicit I mean by these and their related terms.\autocites[The topic of \psy has received a large amount of very sophisticated attention in recent years. See especially][]{mroczek2016}{tigchelaar_tigchelaar2014}{reed_towsend-moulie2011}{reed_jts2009}{reed_ditomasso-turcescu2008}{najman_hilhorst-puech2007}{najman2003}

 In the simplest terms, \psa are texts which fictively purport to be written by figures (typically) from the ancient past. For our purposes, I would like to further distinguish between texts which \emph{portray themselves} and texts which were latter \emph{attributed to} ancient figures. Bernstein helpfully distinguishes between these two phenomena by labeling the former ``authoritative'' \psy and the latter ``decorative'' \psy.\footnote{He also identifies a third form, ``convenient'' \psy which is located somewhere between the two. I do not find this category as helpful. \autocite[3--7]{bernstein_chazon-etal1999}.} While the two phenomena are no doubt related, it is the act of writing in the name of another figure which interests me. Thus, Ps 23, again, although attributed to David, I assume was not \emph{actually} written by him, nor was it written \emph{as if} it had been written by him. Thus, Ps 23 should probably be considered ``decorative'' \psy. Major portions of 1 Enoch, on the other hand (in particular the latter three books, Astronomical Writings [72--82], Dream Visions [83--90], and the Epistle of Enoch [91--107]), were \emph{written as though} they were written by Enoch himself, and therefore fall into the category of ``authoritative'' \psy. Less clear-cut examples, however, require a more nuanced definition. For example, Deuteronomy is not generally referred to as among the \psa (see below), yet, from a literary perspective, it is framed as \emph{had-dəbārîm ʾăšer dibber mōšeh ʾel-kol-yiśrāʾēl} ``the words which Moses spoke to all Israel'' (Deut 1:1a). Was Moses the author? Many Jews and Christians from antiquity up to (and for some, including) the modern era, of course, believed so. But whether Deuteronomy was \emph{written} as \psa or just attributed to Moses is difficult to say with certainly. What we \emph{can} say is that there are concrete literary cues within Deuteronomy which suggest Mosaic authorship more strongly than, say Genesis, which was also attributed to Moses in antiquity.

 The ancient use of the term \psa denoted spurious texts which Church leaders believed to be intentionally misleading about their authorship.\footnote{See esp.~Hist. Eccl. 6.12.2 where the Bishop of   Antioch, Serapion, refers to the * Gospel of Peter* among the a number   of works ``falsely attributed'': γάρ, ἀδελφοί, καὶ Πέτρον καὶ τοὺς   ἄλλους ἀποστόλους ἀποδεχόμεθα ὡς Χριστόν, τὰ δὲ ὀνόματι αὐτῶν   ψευδεπίγραφα ὡς ἔμπειροι παραιτούμεθα, γινώσκοντες ὅτι τὰ τοιαῦτα οὐ   παρελάβομεν. ``For we, brothers, accept both Peter and the other   apostles as Christ, but we skillfully reject those falsely ascribed   writings, knowing that they were not handed down to us.''} The number of (esp.~Jewish) \psgraphical texts discovered within the past century provide good reason to question the assumption that pseudonymous authors's intentions were to deceive their readers.\autocites[53--58]{mroczek2016}[See also][]{reed_jts2009} Thus, I wish to eschew the value judgments of this ancient usage. At the other end of the spectrum, in some scholarly discourse, the term ``\psa'' has become generalized to encompass any text written in around the turn of the era which did not make it into the canon of rabbinic Judaism or early Christianity. Bernstein observes, for example, that although the first volume of James Charlesworth's two-volume \emph{Old Testament Pseudepigrapha} contains formally \psgraphic works, the second volume includes many which do not meet the formal definition of \psa.\autocites[2]{bernstein_chazon-etal1999}{charlesworth_OTP} This expansive practice, likewise, is not particularly helpful for clarifying the term and so I will attempt to restrict my useage to a more clearly defined set of criteria.

 Thus I will use the terms \psy and \psa to refer to texts (or practices) which seem to actively push the reader to construct an implied author whose identity would have been well-known and meaningful to its reader and who (typically) would have lived in the distant past.

 The term ``implied author'' also deserves a clear definition; I have adopted that of H. Porter Abbott:

 \begin{quote} Neither the real \emph{author} nor the \emph{narrator}, the implied author is the idea of the author constructed by the reader as she or he reads the \emph{narrative}. In an \emph{intentional reading}, the implied author is that sensibility and moral intelligence that the reader gradually constructs to infer the intended meanings and effects of the narrative. The implied author might as easily (and with greater justice) be called the ``inferred author'' \autocite[235]{abbott2008} \end{quote}

 Because the implied author is a construction of the reader, it is frequently not desirable to talk about this construct as a literary feature so much as a heuristic for intentional reading. Part of the advantage of basing the definition of \psy on the idea of the implied author is that it mitigates any prejudice toward the intention of the author to deceive (maliciously or otherwise) his reader. Thus, the question of ``who'' the implied author \emph{is} generally misses the point. In the case of \psy, however, the central formal characteristic of the work seems to be the \emph{intentional construction of a known implied author}. Therefore, one could make the case that what is characteristic about \psgraphic texts is the intentionality on the part of the real author to shape the processes by which their readers's construct an implied author. Where Abbott notes that the term ``inferred author'' may be a better term than ``implied author'' (the connotation shifts the active role to the reader) when discussing \psy, ``implied'' still fits quite well (since we can attribute some intentionality to the real author). The implied author, therefore, provides an important point of contact between the reader and the (real) author.

Taking the above ideas together, and with respect to the \ga, we can see that the homogenization of the (putative) sources of the \ga brought about by their presence in a ``collection'' (whether truly composite or whether as a literary device of a single author), allows for readers to traverse \ga with what one might consider to be a single implied author.\footnote{This assertion relies on Abbott's description of the Implied author as a ``sensibility and moral intelligence that the reader gradually constructs to infer the intended meanings and effects of the narrative'' (235). Insofar as this is the case, I think it's fair to say that the entirty of \ga can be read with the same ``sensability and moral intellegence,'' regardless of who the narrator is portrayed to be at any given moment.\autocite[235]{abbott2008}.} What we have not addressed, however, is how these literary features mayhave functioned with respect to social memory.

 \subsection{Scripture, \Psy and Memory Construction}

 The vast majority of the Hebrew Bible is narrated in the third-person omniscient and is formally anonymous. There are, of course, exceptions to this generalization, most notably within the prophetic corpus (such as Isa 6--8), the so-called Nehemiah Memoir (Neh 11--13), and perhaps works such as Deuteronomy and Song of Songs. But for the lion's share of the biblical text, the the implied author seems to operate invisibly.

 The rhetorical force of this particular authorial voice, as observed by Erhard Blum, is significant for the function of the Hebrew Bible's participation in the collective memory of the communities that claim it as their own. Although the implied author does occasionally engage directly with the reader by offering explanatory observations (for example where the author inserts phrases like ``this is why\ldots{}'' or ``\ldots{}until this day''), for all intents and purposes, the implied author presents as both \emph{reliable} and \emph{authoritative} without a hint of subjectivity. As Blum puts it, ``In this sense the narrative does not distinguish the depiction from the depicted.''\autocite[33]{blum_barton-etal2007} Put another way, the text does not acknowledge that it \emph{has} an author, it simply \emph{is}. The rhetorical effect of this invisible, omniscient author is to collapse the knowledge gap between the reader and the events narrated by removing the author from view. This move, according to Blum, allows the text to convey ``an unmediated truth claim which is not based on the author's distinguishable critical judgments and convictions.''\autocite[33]{blum_barton-etal2007} The effectiveness of this implied author, according to Blum, is tied to the pragmatics of the text, that is, tied to the context of the biblical narratives as scripture (though, Blum does not refer to ``scripture'' \emph{per se}). The implied audience of the biblical narratives by-and-large can be understood as group-insiders for whom the biblical text worked to reinforce group identity. Such ``unmediated truth claims'' \emph{were} in fact mediated and reinforced by those who (orally or otherwise) transmitted the tradition from one generation to another.\autocite[33]{blum_barton-etal2007} In other words, one might say that the implied author of the biblical text is the community's collective memory.\autocite[Blum writes, ``If we assume that the traditional literature was primarily transmitted through oral means, than the narrator who is speaking supplies the material with a personal presence; he is not present as an author who judges and evaluates his sources from a critical distance, but as a `transmitter' who participates in the tradition itself and is able to lend it credence through his own personality, his standing, and/or his office.''][33]{blum_barton-etal2007}

 In contrast to the omniscient implied biblical author, the \ga frames itself as a collection of first-person accounts which formally fall into the category of \psy. If we take seriously Blum's characterization of the way that the biblical text may have engaged with the collective memory of Israel based on the formal, narratological features of the text, it stands to reason that the \ga as first-person \psy would engage that collective memory in a different way despite the relative similarity of the textual content. The \psgraphic quality of \ga shapes the way that the text engages with the remembered past by describing the biblical story through the mouthes of important figures. Here ``story'' refers to the abstract sequence of actions which the narrative describes. The \emph{way} a story is recounted, on the other hand, is referred to by narratologists as \emph{narrative discourse.} Thus the \ga's change from third-person omniscient to a \psgraphical first-person narrative can be understood as a change in \emph{narrative discourse} which, broadly, retains the same \emph{story} as that of the biblical text.

 Approaching these questions from the perspective of Social Memory Studies asks us to think about the way that differing social frameworks and cognitive contexts may have allowed for or demanded presenting this material in a different form than its \emph{Vorlage}.

 One of the difficulties in dealing with \psy is the apparently divergent ways that the authors and original readers may have understood \psy as compared to the way that later groups (e.g., Church Fathers, modern scholars) treat it. The crux of the issue, it seems to me, is less to do with whether the author intended to ``deceive'' his audience, and more to do with whether the readers understood themselves to be reading something ``authentic'' or were willing participants in an authorial fiction. Yet, even language of ``authenticity'' or ``fiction'' presupposes that such terms were a meaningful part of the discourse surrounding ``scripture'' during the late \secondtemple period.

 \section{Abram in the Diaspora: Keying and Framing in \GA}
 While retelling portions of Genesis as first-person narrative reorients the way that the story engages with the received tradition and collective memory at the macro-level the narrative of the \ga is not simply a straight-forward retelling of Genesis from the perspectives of Lamech, Noah, and Abram. Indeed, what is most compelling about \rwb texts very often is the ways that they depart from the biblical narrative. These departures can come at the level of story by adding, removing, or rearranging events or at the level of narrative discourse by describing events differently or with different emphases. In the case of \ga, and in particular in the account of Abram in cols. 19--22, the biblical narrative has been recast as a (first-person) Hellenistic novella in a similar vein to other well-known Second Temple Jewish works such as the narrative portions of Daniel (including the Greek additions), Esther, Tobit, and (arguably) the so-called Joseph novella of Genesis 37 and 39--50.\autocites*[See especially Lawrence Wills work on the Jewish novels and novellas in antiquity:][]{wills2002}[as well as his important earlier works:][]{wills1995}[and][]{wills1990}

 The reading of \ga 19--20 as a Hellenistic Jewish novella has recently been very thoroughly explicated by Blake Jurgens, who has further argued that the utilization of Hellenistic literary motifs and structures in \ga altered the overall purpose the the pericope for the purpose of edifying Jews living in the Hellenistic world in the shadow of empire.\autocite{jurgens_jsj2018} Although much of Jurgen's paper is based on long-established observations about the literary influences on \ga, he makes the important discursive turn toward the audience by claiming that the \ga was meant to be useful to readers:

 \begin{quote} By imbuing its story with literary tropes and techniques similar to those found in Dan 1--6, Esther, and other Jewish texts arising out of the Hellenistic period, the author successfully attends to the narratival ambiguities of Gen 12:10--20 through interpretive expansion upon the latent exegetical links of the text while concurrently modifying the narrative to appeal to contemporary literary expectations.\autocite[27]{jurgens_jsj2018} \end{quote}

 The process of this transcription, which he terms ``fictionalization,'' is described by Jurgens in the six distinct narrative units within cols. 19-20 of \ga which describe Abram and Sarai's sojourn in Egypt: the decent into Egypt, Abram's dream vision, the banquet Scene, praise of Sarai's beauty, Abram's prayer, and the final court contest. In each section Jurgens notes the ways that the \ga utilizes literary structures common to it broader Hellenistic mileu to rewrite the the events of this story. Jurgens offers a very thorough description of the ways that the \ga utilizes these literary structures and makes a plausible claim that these changes were meant to engage readers in familiar style. Thinking in terms of social memory, however, we can appreciate the way that the story of Abram and Sarai in Egypt is ``remembered into'' the social context of Hellenistic Judaism and is fitted into the contemporary social frameworks (read: literary conventions) therein.

 While \halbwachs addressed the fact that memories are shaped by the present, recently Barry Schwartz has attempted to more clearly articulate this process. One of the most important contributions of Schwartz's work in this area is his conviction that the interactions between the remembered past and the present are not unidirectional. Where \halbwachs limits his discussion to the describing the ways that memories of the past are shaped by the present, Schwartz sees the past as a potent force in the present as well. In other words, not only does the present influence the way the past is remembered, but the past \emph{itself} (that is, both the remembered and ``actual'' past) also effects the present.

 Schwartz employs two terms, ``keying'' and ``framing,'' to describe this additional dimension to the way that the past impacts the present. On the one hand the idea of ``keying'' can be understood as way of


 \subsection{Setting}

 One of the primary features of these novellas is their setting. Jurgens notes that, typically, these Jewish novellas are set in the diaspora, which invariably place the Jewish (or, in Tobit and Judith's case, Israelite) protagonist under the hegemony of a foreign power. In the case of \ga, although not properly ``diaspora,'' Abram is a sojourner in a foreign land and is under foreign hegemony. Moreover, from a modern perspective, these stories have a tendency to commit rather egregious factual errors about certain historical particulars such as the names of rulers (Judith 1:1; Dan 4; Tobit) and geographic items (Tobit 5:6). Likewise, \ga seems to utilize details which almost certainly were inventions of the author (or an earlier tradant) such as referring to ``Pharaoh Zoan'' (we know of no such figure) and Herqanos, a name popular in the Ptolemaic period, but not attested otherwise as well as referring to the ``Karmon River'' (probably the Kharma canal), as the one of the seven heads of the Nile river, which it is not.\autocites[7]{jurgens_jsj2018}[See also][50--59]{machiela_as2010}[197--199]{fitzmyer2004} These details, according to Jurgens, are meant to create a sense of verisimilitude and authenticity within the narrative. Thus, although the story of Abram's sojourn in Egypt as narrated in the biblical text engages with discourses of the \emph{foundation} of Israel, the narrative of the \ga seems to be turning the story to engage with the contemporary discourses around the idea of \emph{diaspora}.


 \subsection{Abram in the Court of a Foreign King}
 \subsection{Abram the Sage}
 \subsection{Abram the Oracle}
 \section{Inner-biblical Interpretation}