% !TeX root = ../dissertation.tex

\begin{aramaictext}
    \versenum{XIX 10}
    \dots
    \vacat
    והוא כפנא בארעא ד̇א כ̇ולא̇ ושמעת די ע[בו]ר֯א 
    ה֯[וא] במצרין ונגדת
    \versenum{11}
    ל̇[מעל] לאר̇̇ע̇ מ̇צרין [\hspace*{1.5em}]◦◦◦[\hspace*{1.5em}עד די דבק]ת לכ̇רמונא נהרא חד מן 
    \versenum{12}
    רא֯ש֯י֯ נה̇ר֯א֯ ◦◦◦[\hspace*{1.5em}]עד כען אנחנא בגו א֯ר֯ע֯נ֯א֯ [וח]לפת שבעת ראשי נהרא דן 
    די
    \versenum{13}
    ◦◦◦[\hspace*{1.5em}]א כען חלפנא א̇ר̇ענא ועלנא לארע בני חם לארע מצרין
\end{aramaictext}

\begin{translation}
    \versenum{10}
    Now, there was famine in all of that land, but I heard that there was grain in Egypt. So I proceeded 
    \versenum{11}
    to enter into the land of Egypt [\hspace*{1em}] [until I reached] the Carmon river, one of 
    \versenum{12}
    the heads of the river [\hspace*{1em}] unto now we had been within the boundaries of our (own) land. And I passed by the seven heads of the Nile river which [
    \versenum{13}
    ] now, we had crossed our land and we entered into the land of the children of Ham, into the land of Egypt.
\end{translation}

\begin{aramaictext}
\versenum{14}
    \vacat
    וחלמת אנה אברם חלם בלילה מעלי לאר̇ע̇ מ̇צ̇ר֯י֯ן וחזית בחלמי֯ [וה]א 
    ארז חד ותמרא
    \versenum{15}
    כחדא צמ[חו] מן ש֯ר֯[ש חד] ובני אנוש אתו ובעון למק̇ץ ולמ̇עקר ל[א]ר̇זא 
    ולמ̇ש̇ב̇ק̇ ת̇מ̇ר̇\textsuperscript{ת}א בלחודיהה
    \versenum{16}
    ואכליאת תמרתא ואמרת אל תקוצ̇ו ל[א]רזא ארי תרינא מן שרש [ח]ד֯ 
    צ֯מ֯ח֯[נ]א ושביק ארזא בטלל תמרתא
    \versenum{17}
    ולא [אתקץ] \vacat
\end{aramaictext}

\begin{translation}
    \versenum{14}
    And I, Abram, dreamed a dream in the night (concerning) my entry to the land of Egypt.  And I saw in my dream a cedar and a date palm
    \versenum{15}
    sprouting as one from one root. And people approached and sought to cut down and to uproot the cedar so as to leave the date palm alone.
    \versenum{16}
    But the date palm shouted and said, ``do not cut down the cedar because the two of us sprouted from one root.'' And the cedar was left alone with the help of the date palm
    \versenum{17}
    and it was not [cut down].
\end{translation}

\begin{aramaictext}
    ואת̇עירת בליליא מן שנתי ואמרת לשרי אנתתי חלם 
    \versenum{18}
    חלמת [אנה וא]דחל [מן] ח֯למ̇̇א֯ דן ואמרת לי א̇שתעי לי חלמך ואנדע ושרית 
    לאשתעיא לה חלמא דן
    \versenum{19}
    [וחוית] ל[ה פשר] חלמא [דן ו]אמ[רת] ◦◦◦ די יבעון למקטלני ולכי למשבק
    [ב]ר̇ם דא כול טבותא
    \versenum{20}
    [די תעבדין עמי] בכול א֯ת֯ר֯ די [נהך לה אמרי] עלי די אחי הוא ואחי בטליכי
    ותפלט̇ נפשי בדיליכי
    \versenum{21}
    [\hspace*{1.5em}יבעון] לאע[ד]יותכי מני ולמקטלני ובכת שרי ע̇ל מ̇לי בליליא דן
    \versenum{22}
    [\hspace*{1.5em}]◦◦◦[\hspace{8em}] ופרעו צ[ען\hspace*{1.5em}] שרי למפ̇נה לצען
    \versenum{23}
    [עמי והסתמרת י]ת̇ירא בנפשה די לא יחזנה כול [אנש חמש שני]ן ולסוף
    חמש שניא אלן
    \versenum{24}
    [\hspace*{1.5em}אתו] תלתת גברין מן רברבי מצרי[ן\hspace*{0.75em}] די פרע[ו] צע[ן] על מל[י] ועל
    אנתתי והווא יהבין
    \versenum{25}
    [לי מתנן שגיאן ובעו] ל[י] ל[אודעא] טבתא וחכמתא וקושטא וקרית קודמיהון
    ל[כתב] מלי חנוך
    \versenum{26}
    [\hspace{8em}] בכפנא די [\hspace*{1.5em}]◦◦◦ ולא ◦◦◦ין למקם עד די ◦◦◦ מלי
    \versenum{27}
    [\hspace{2em}]ל[\hspace*{1.5em}]במאכל שגי ובמשתה[\hspace{8em}]חמרא
    \hspace*{1.5em} 
    \versenum{28}
    [\hspace{3em}]
\end{aramaictext}

\begin{translation}
    Then I awoke in the night from my sleep and I said to Sarai my wife, ``I dreamed
    \versenum{18}
    a dream and I am afraid on account of this dream.'' And she said to me, ``tell me your dream so that I may know (it).'' And I began to recount to her this dream 
    \versenum{19}
    [and I showed her the interpretation] of this dream. And I [said \hspace*{1em}] who will seek to kill me, but you, to leave (you) alone. But this is the good thing 
    \versenum{20}
    [that you must do for me] in every place that [we might go, say] about me ``he is my brother'' so that I might live because of you and (so that) my life might be rescued because of you. 
    \versenum{21}
    [they will seek] to take you away from me and to kill me. And Sarai wept on account of my words that night.
    \versenum{22}
    [\hspace*{1em}] [\hspace*{1em}] and Pharoah [Zoan \hspace{1em}] Sarai to turn to Zoan
    \versenum{23}
    [with me and she] was very [conscientious] lest any [man] see her [for five years]. Yet, when these five years had passed
    \versenum{24}
    [There came] three men from the princes of [Egypt] of Pharoah Zoan on account of my words and on account of my wife. And they gave
    \versenum{25}
    [me many gifts. And they asked] me for [knowledge] of goodness and wisdom and righteousness. So I read them the book of the words of Enoch.
    \versenum{26}
    [\hspace*{1em}] in the famine which [\hspace*{1em}] to stand until [\hspace*{1em}] the words of
    \versenum{27}
    [\hspace*{1em}] with much eating and drinking[\hspace*{1em}] wine.
    \versenum{28}
    [\hspace*{1em}]
\end{translation}