% !TEX root = dissertation.tex
\begin{ethiopictext}
    \versenum{Jubilees 1:9}
    እስመ~፡ ይረስዑ~፡ ኵሎ~፡ ትእዛዝየ~፡ 
    ኵሎ~፡ ዘአነ~፡ እኤዝዞሙ~፡ ወየሐውሩ~፡ ድኅረ~፡ አሕዛብ~፡ ወድኅረ~፡
    ርኵሶሙ~፡ ወድኅረ~፡ ኀሳሮሙ~፡ ወይትቀነዩ~፡ ለአማልክቲሆሙ~፡ 
    ወይከውንዎሙ~፡ ማዕቀፈ~፡ ወለምንዳቤ~፡ ወለፃዕር~፡ ወለመሥገርት~፡
    \versenum{10}
    ወይትሐጐሉ~፡ ብዙኃን~፡ ወይትአኀዙ~፡ ወይወድቁ~፡ ውስተ~፡
    እደ~፡ ፀር~፡ እስመ~፡ ኀደጉ~፡ ሥርዓትየ~፡ ወትእዛዝየ~፡ ወበዓላተ~፡
    ኪዳንየ~፡ ወሰንበታትየ~፡ ወቅድሳትየ~፡ ዘቀደስኩ~፡ ሊተ~፡ በማእከሎሙ~።
    ወደብተራየ~፡ ወመቅደስየ~፡ ዘቀደስኩ~፡ ሊተ~፡ በማእከለ~፡
    ምድር~፡ ከመ~፡ እሢም~፡ ስምየ~፡ ሳዕሌሁ~፡ ወይኅድር~።
    \versenum{11}
    ወገብሩ~፡
    ሎሙ~፡ ፍሥሐታተ~፡ ወኦመ~፡ ወግልፎ : ወሰገዱ~፡ ዘዘ~፡ ዚአሆሙ~፡ 
    ለስሒት~፡ ወይዘብሑ~፡ ውሉዶሙ~፡ ለአጋንንት~፡ ወለኵሉ~፡ ግብረ~፡
    ስሕተተ~፡ ልቦሙ~።
    \versenum{12}
    ወእፌኑ~፡ ኀቤሆሙ~፡ ሰማዕተ~፡ ከመ~፡
    አስምዕ~፡ ሎሙ~፡ ወኢይሰምዑ~፡ ወሰምዕተኒ~፡ ይቀትሉ~፡ ወለእለሂ~፡
    የኀሥሡ~፡ ሕገ~፡ ይሰድድዎሙ~፡ ወኵሎ~፡ ያፀርዑ~፡ ወይዌጥኑ~፡ ለገቢረ~፡
    እኩይ~፡ በቅድመ~፡ አዕይንትየ~፡
    \versenum{13}
    ወአኀብእ~፡ ገጽየ~፡
    እምኔሆሙ~፡ ወእሜጥዎሙ~፡ ውስተ~፡ እደ~፡ አሕዛብ~፡ ለፂዋዌ~፡
    ወለሕብል~፡ ወለተበልዖ~። ወአሴስሎሙ~፡ እማእከለ~፡ ምድር~፡
    ወእዘርዎሙ~፡ ማእከለ~፡ አሕዛብ~፡
    \versenum{14}
    ወይረስዑ~፡ ኵሎ~፡ ሕግየ~፡
    ወኵሎ~፡ ትእዛዝየ~፡ ወኵሎ~፡ ፍትሕየ~፡ ወይስሕቱ~፡ ሠርቀ~፡ ወሰንበተ~፡
    ወበዓለ~፡ ወኢዮቤለ~፡ ወሥርዓተ~።
\end{ethiopictext}

\begin{transliteration}
    \versenum{Jubilees 1:9}
    ʔəsma yərassəʕu \kw{ə}llo təʔzāzəya
    % yərassəʕu         g impf 3mp √rsʕ 'to forget' Les. 473
    % təʔzāzya          təʔzāz + 1cs 'law'
    \kw{ə}llo za-ʔana ʔəʔēzzəzomu wa-yaḥawwəru dəḫra ʔaḥzāb wa-dəḫra
    % ʔəʔēzzezomu       d impf 1cs +3mp √ ʔzz 'to command' Les. 53
    % yaḥawwəru         g impf 3mp √ḥwr 'to go' Les. 249
    % dəḫra             'back, past, after, then' Les. 129
    % ʔaḥzāb            pl. of ḥəzb 'nation, people, tribe' Les. 253
    rə\kw{ə}somu wa-dəḫra ḫasāromu wa-yətqan\-nayu la-ʔamāləktihomu
    % rəkwsomu          n. √rkws 'to be unclean, impure' Les. 470
    % ḫasārom           n. √ḫsr 'dishonor' Les. 265
    % yətqannayu        tG impf. 3mp 'serve' √qny Les. 437
    % ʕamālektihomu     pl. of ʔamlāk + 3mp '(false) gods, idols' Les. 344
    wa-yəkawwənəwwomu māʕəqafa wa-la-məndābē wa-la-ḍāʕr [ṣāʕr] wa-la-maśgart
    % yəkawwənəwwomu    g impf. 3mp (ənu- > ənəww- Lambd. 64) √kwn 'become'
    % māʕəqafa          √ʕqf 'to trip up' Les 67
    % məndābē           'tribulation, affliction' Les. 348
    % ḍāʕər             √ṣʕr 'anguish, trouble' Les. 544
    % maśgart           √śgr II 'snare, trap' Les 527
    \versenum{10}
    wa-yətḥag\gw{a}lu [yəthag\gw{a}lu] bəzuḫān wa-yətʔaḫḫazu wa-yəwaddəqu wəsta
    % yətḥagolu         tG impf √hgwl 'destroy'
    % bəzaḫān           'many' √bzḫ 'be numerous' Les 117
    % yətʔaḫḫazu        tG impf. 3mp √ʔḫz 'take, catch' Les. 14
    % yəwaddəqu         G impf 3mp √wdq 'collapse, go to ruin' Les.
    ʔəda ḍarr ʔəsma ḫadagu śərʕātəya wa-təʔzāzəya wa-baʕālāta
    % ʔəda              'hand' pl. ʔədaw Les. 7
    % ḍar               'enemy' pl. ʔaḍrār Les. 152
    % ḫadagu             g pf. 3mp √ḫdg 'to abandon' Les. 258
    % śərʕāteya          pl. 'ordinances' √śrʕ Les. 533
    % baʕālāta          pl. 'festivals' Les. 83
    kidānǝya wa-sanbatātǝya wa-qəddǝsātǝya za-qaddasku lita ba-māʔkalomu
    % kidāneya           'covenant' √kyd Les 301
    % sanbatāteya        'sabbaths'
    % qəddesāteya        'holy things' Les 422
    % qaddasku           d pf. 1cs √qds 'to make holy' Les. 422
    % ba-māʔkal          'amongst'
    wa-dabtarāya wa-maqdasǝya za-qaddasku lita ba-māʔkala
    % dabtarāya         'tabernacle'
    % maqdasya          'temple'
    % qadasku           g pf. 1cs √qds 'to make holy' Les. 422
    mədr kama ʔəśim səməya lāʕlēhu wa-yəḫdər
    % ʔəśim         g. subj. 1cs √śym 'put, place' Les. 539
    % səmya         'name' +1cs Les. 504
    % lāʕlēhu       prep. 'upon'
    % yəḫdər        g subj 3ms √ḫdr 'to dwell'  Les. 258
    \versenum{11}
    wa-gabru
    % gabru             g pf. 3mp √gbr 'to make'
    lomu fəśḥatāta wa-ʔoma [ʕoma] wa-gəlfo wa-sagadu zazza ziʔahomu
    % fəśḥatāta         pl. 'high place' √fśḥ II Les. 168
    % ʔoma              ʕoma 'grove' Les. 62
    % gəlfo             adj. 'carved' √glf 'to carve' Les. 190
    % sagadu            g pf 3mp √sgd 'bow down, prostrate'
    % zaza              each
    % ziʔahomu          each of them (probably zazazaiʔahomu)
    la-səḥit wa-yəzabbəḥu wəludomu la-ʔəgānənt wa-la-\kw{ə}llu gəbra
    % səḥit             'sin, error' √sḥt 'to stray, err' Les. 494
    % yəzabbəḥu         g impf 3mp √zbḥ 'to sacrifice' Les. 631
    % ʔəgānənt          'demons' sn. gānən Les 198
    % gəbra             'thing' √gbr 'to do, make' Les 178
    səḥtata ləbbomu
    % saḥtata           'sinful' √sḥt 'to stray, err' Les. 494
    % ləbbomu           'hear, mind'
    \versenum{12}
    wa-ʔəfēnnu ḫabēhomu samāʕta kama
    % ʔəfēnnu           d. impf 1cs √fnw 'to send' Les. 163
    % ḫabē-             'toward, near, to'          
    % samāʕta           'witnessses' samāʕi √smʕ 'hear'
    ʔasməʕ [ʔāsməʕ] lomu wa-ʔiyyəsamməʕu  wa-samāʕta-ni yəqattəlu wa-la-ʔəlla-hi
    % ʔasməʕ            CG Subj 1cs √smʕ 'to bear witness'
    % ʔiyyəsammeʕu      g impf 3mp √smʕ 'to listen' Les. 501
    % samāʕta           'witnessses' samāʕi √smʕ 'hear' 
    yaḫaśśəśu ḥəgga yəsaddədəwwomu wa-\kw{ə}llo yāḍarrəʕu wa-yəwēṭṭənu la-gabira
    % yaḫaśśəśu         g impf √ḫśś 'to seek' Les. 266
    % yəsaddədəwwomu    g impf 3mp √sdd 'persecute, drive away'
    % yāḍarrəʕu         CG impf 3mp √ḍrʕ 'to annul, leave aside'
    % yəwēṭṭənu         D impf 3mp √wṭn Les. 623
    % gabira            G inf √gbr 'to do'
    ʔəkuya ba-qədma ʔaʕəyyəntəya
    % ʔəkuya            'evil'
    % ba-qədmu          'before'
    % ʔaʕəyyəntəya      'eyes' sn. ʕayn +1cs Les. 79
    \versenum{13}
    wa-ʔaḫabbəʔ gaṣṣəya
    % ʔaḫabbəʔ          g impf 1cs √ḫbʔ 'hide, conceal' Les. 253
    % gaṣṣəya           'face' Les. 205
    ʔəmmənēhomu wa-ʔəmēṭṭəwomu wəsta ʔəda ʔaḥzāb la-ḍiwāwē
    % ʔəmēṭəwwomu       d impf 1cs √mṭw 'hand over, deliver' Les. 374
    % ʔaḥzāb            pl. of ḥəzb 'nation, people, tribe' Les. 253
    % ḍiwāwē            'captivity' √ḍww 'take prisoner' Les. 153         
    wa-la-ḥəbl wa-la-tabalʕā wa-ʔasēssəlomu [ʔāsēssəlomu] ʔəm-māʔkala mədr
    % ḥəbl              'spoils' Les. 223
    % tabalʕā           tG pf 3fs? 'devouring' √blʕ 'to eat' Les. 94
    % ʔasēssəlomu       CD impf 1cs +3mp 'remove' Les. 516
    wa-ʔəzarrəwomu māʔəkla ʔaḥzāb
    % ʔəzarrəwomu       G impf 1cs +3mp √zrw 'scatter, disperse' Les. 644
    % ʔaḥzāb            pl. of ḥəzb 'nation, people, tribe' Les. 253
    \versenum{14}
    wa-yərassəʕu \kw{ə}llo ḥəggəya
    % yərassəʕu         g impf 3mp √rsʕ 'to forget' Les. 473
    wa-\kw{ə}llo təʔzāzəya wa-\kw{ə}llo fətḥəya wa-yəsəḥḥətu śarqa wa-sanbata
    % təʔzāzya          təʔzāz + 1cs 'law'
    % fətḥəya           fətḥ 'precept' √ftḥ Les. 170
    % yəsəḥḥətu         G impf 3mp √sḥt 'to err, make a mistake' Les 494
    % śarqa             'rising' here 'beginning of the month' √śrq Les. 534
    % sanbata           'sabbath' Les. 505
    wa-baʕāla wa-ʔiyyobēla wa-śərʕāta
    % śərʕāta           'ordinance' √śrʕ Les. 533
\end{transliteration}

\begin{translation}
    \versenum{Jubilees 1:9}
    Therefore they will forget my law---%
    all that I am commanding them---and they will go after the nations and after
    their impurity and after their dishonor. And they will serve their (false) gods
    and they will become a hindrance and an affliction and a torment and a snare for them.
    \versenum{10}
    Many will be destroyed and they will be captured and will fall into
    the hand of the enemy because they abandoned my ordinances and my laws and festivals of 
    my covenant and my sabbaths and my sacred items which I sanctified for myself among them.
    Also my tabernacle and my \temple which I sanctified for myself in midst of
    the land that I might establish my name upon it and it might dwell (there).
    \versenum{11}
    And they made
    for themselves high places and a grove, and a carved image and they bowed down, each one of them,
    to error. And they will sacrifice their children to demons and to every thing
    of their sinful mind.
    \versenum{12}
    And I will send witnesses to them that
    I may testify to them, but they will not listen and (instead) they will kill the witnesses and even
    those (who) seek after the law, they will persecute. And they will leave aside everything and they will begin to do
    evil before my eyes.
    \versenum{13}
    And I will hide my face
    from them and I will deliver them into the hand of the nations for captivity,
    for spoils, and for being devoured. I will remove them from the midst of the land
    and I will scatter them among the nations.
    \versenum{14}
    They will forget my whole law,
    and all my commandments, and all my precepts. And they will err regarding the new moon and the sabbath
    and the festival and the jubilee and the ordinances.
\end{translation}