% !TeX root = ../dissertation.tex

In his seminal work \citetitle*{vermes1961}, \Vermes introduced the term ``\rwB'' into the discussion of \secondtemple Jewish literature as part of a larger project to trace the development of haggadic traditions from the late \secondtemple period into the rabbinic period. \vermes used the term \rwB to describe a number of texts which closely follow portions of the biblical narrative but also augment, elide, and emend the text in ways which produced new literary works in their own right. According to \vermes, through this exegetical process, ``the midrashist inserts haggadic development into the biblical narrative'' in order to ``anticipate questions, and to solve problems in advance.''\autocites[95]{vermes1961}[see also][]{vermes_zsengeller2014} \vermes traced these interpretive traditions historically and attempted to demonstrate sn interpretive continuity between the \secondtemple and rabbinic periods. Although the formal characteristics of these narratives differed from later midrash, \rwB texts displayed the same kinds of ``midrashic'' tendencies. In \vermes' conception, therefore, the authors of \rwB texts \emph{implicitly} made use of interpretive traditions that later works such as the Talmud and Mishnah expressed \emph{explicitly}. 

Since the publication of \citetitle{vermes1961}, \vermes's concept of \rwB has taken on a life of its own and developed into its own discreet area of study as scholars from various related disciplines have reused, reinterpreted, and redefined the term. The scope and nuance of of the term \rwB has shifted a great deal in the intervening years, yet, the trajectory set by \vermes nearly sixty years ago has remained reasonably consistent by focusing on the relationships that exist between these unique works and the scriptural texts that inspired them. 

In this chapter I will trace the emergence and evolution of the concept of \rwB from \vermes's use in \citetitle{vermes1961} to the present focusing on three key questions and ideas which have shaped the scholarly discourse around \rwB studies: 1) the terminology surrounding \rwB, 2) what works should fall under the rubric of \rwB? and 3) whether \rwB constitutes a literary genre, a process, or some combination of the two. Taking these questions into account, I will then offer my own suggestions on how the \rwb conversation can be augmented by memory studies. [[TODO: Expand this a bit]]