% !TEX root = dissertation.tex
\begin{ethiopictext}
    \versenum{Jubilees 1:9}
    እስመ ፡ ይረስዑ ፡ ኵሎ ፡ ትእዛዝየ ፡ 
    ኵሎ ፡ ዘአነ ፡ እኤዝዞሙ ፡ ወየሐውሩ ፡ ድኅረ ፡ አሕዛብ ፡ ወድንረ ፡
    ርኵሶሙ ፡ ወድንረ ፡ ኀሳሮም ፡ ወይትቀነዩ ፡ ለዐማልክቲሆሙ ፡ 
    ወይከውንዎሙ ፡ ማዕቀፈ ፡ ወለምንዳቤ ፡ ወለፃዕር ፡ አውለመሥገርት ፡
    \versenum{10}
    ወይትሐጐሉ ፡ ብዙኃን ፡ ኣአይትአኀዙ ፡ ወይወድቁ ፡ ውስተ ፡
    እደ ፡ ፀር ፡ ኣመ ፡ ኀደጉ ፡ ሥርዓትየ ፡ ወትእዛዝየ ፡ ወበዓላተ ፡
    ኪዳንረ ፡ ወሰንበታትየ ፡ ወቅድሳትየ ፡ ዘቀደስኩ ፡ ሊተ ፡ በማእከሎሙ ።
    ወደብተራየ ፡ ወመቅደስየ ፡ ዘቀደስኩ ፡ ሊተ ፡ በማእከለ ፡
    ምድር ፡ ከመ ፡ እሢም ፡ ስምየ ፡ ሳዕሌሁ ፡ ወይንድር ።
    \versenum{11}
    ወገብሩ ፡
    ሎሙ ፡ ፍሥሐታተ ፡ ወኦመ ፡ ዋግልፎ : ወሰገዱ ፡ ሰሰ ፡ ዚአሆሙ ፡ 
    ለስሒት ፡ ወይዘብሑ ፡ ውሉዶሙ ፡ ለአጋንንት ፡ ወለኵሉ ፡ ግብረ ፡
    ስሕተተ ፡ ልቦሙ ።
    \versenum{12}
    ወእፌኑ ፡ ኀቤሆሙ ፡ ሰማዕተ ፡ ከመ ፡
    አስምዕ ፡ ሎሙ ፡ ወኢይሰምዑ ፡ ወሰምዕተኒ ፡ ይቀትሉ ፡ ወለእለሂ ፡
    የኀሥሡ ፡ ሕገ ፡ ይሰድድዎሙ ፡ ወኵሎ ፡ ያፀርዑ ፡ ወይዌጥኑ ፡ ለገቢረ ፡
    እኩይ ፡ በቅድመ ፡ አዕይንትየ ፡
    \versenum{13}
    ወአኀብእ ፡ ገጽየ ፡
    እምኔቦሙ ፡ ወእሜጥዎሙ ፡ ውስተ ፡ እደ ፡ አሕዛብ ፡ ለፂዋዌ ፡
    ወለሕብሌ ፡ ወለተበልዖ ። ወአኣኤኣሎሙ ፡ እማእከለ ፡ ምድር ፡
    ወእዘርዎሙ ፡ ማእከለ ፡ አሕዛብ ፡
    \versenum{14}
    ወይረስዑ ፡ ኵሎ ፡ ሕግየ ፡
    ወኵሎ ፡ ትእዛዝየ ፡ ወኵሎ ፡ ፍትሕየ ፡ ወይስሕቱ ፡ ሠርቐ ፡ ወሰንበተ ፡
    ወበዓለ ፡ ወኢዮቤለ ፡ ወሥርዓተ ።
\end{ethiopictext}
    \versenum{Jubilees 1:9}

    \versenum{10}

    \versenum{11}

    \versenum{12}
\begin{transliteration}
\end{transliteration}

\begin{translation}
    \versenum{Jubilees 1:9}
    VanderKam: For they will forget all my commandments---everything that I command them---and will follow the nations, their impurities, and their shame. They will serve their gods and (this) will prove an obstacle for them---an affliction, a pain, and a trap.
    \versenum{10}
    Many will be destroyed. They will be captured and will fall into the enemy's control because they abandoned my statutes, my commandments, my covenantal festivals, my sabbaths, my holy things which I have hallowed for myself among them, my tabernacle, and my temple which I sanctified for myself in the middle of the land so that I could set my name on it and that it could live (there)
    \versenum{11}
    They  made for themselves high places, (sacred) groves, and carved images; each of them prostrated himself before his own in  order to go astray. They will sacrifice children to demons and to every product (conceived by) their erring minds.
    \versenum{12}
    I will send witnesses to them so that I may testify to them, but they will not listen and will kill the witnesses. they will persecute those too who study the law diligently. They will abrogate everything and will begin to do evil in my presence.

    \emph{**VanderKam's translation (in the interest of time). I will provide my own text and translation in subsequent drafts}
\end{translation}