% !TeX root = ../dissertation.tex
\begin{aramaictext}
    \versenum{XX 1}
    ◦◦◦[\hspace{3em}]◦◦◦
    \versenum{2}
    ◦◦◦[\hspace{9em}] כמ̇ה ◦◦◦ ושפיר לה צלם אנפיהא וכמא
    \versenum{3}
    [נ]ע֯י֯ם̇ ו֯כ֯מ֯א֯ רקיק לה שער ראישה כמא יאין להון לה עיניהא ומא רגג הוא
    לה אנפהא וכול נץ
    \versenum{4}
    אנפיהא ◦◦◦ כמא יאא לה הדיה וכמא שפיר לה כול לבנהא דרעיהא מא
    שפירן וידיהא כמא
    \versenum{5}
    כלילן וחמיד כול מחזה יד̇[י]הא כמא יאין כפיהא ומא אריכן וקטינן כול
    אצבעת ידיהא רגליהא
    \versenum{6}
    כמא שפירן וכמא שלמא להן לה שקיהא וכל בתולן וכלאן די יעלן לגנון
    לא ישפרן מנהא ועל כול
    \versenum{7}
    נשין שופר שפר̇ה ועליא שפרהא לעלא מן כולהן ועם כול שפרא דן חכמא
    שגיא עמהא ודלידיהא
    \versenum{8}
    יאא וכדי שמע מלכא מלי חרקנוש ומלי תרין חברוהי די פם חד תלתהון
    ממללין שגי רחמה ושלח
    \versenum{9}
    לעובע דברהא וחזהא ואתמה על כול שפרהא ונסבהא לה לאנתא ובעא
    למקטלני ואמרת שרי
    \versenum{10}
    למלכא דאחי הוא כדי הוית מתגר על דילהא ושביקת אנה אברם בדילהא
    ולא קטילת ובכית אנה
    \versenum{11}
    אברם בכי תקיף אנה ולוט בר אחי עמי בליליא כדי דבירת מני שרי באונס
    \vacat
\end{aramaictext}

\begin{translation}
    \versenum{XX 1}
    [\hspace{1em}]
    \versenum{2}
    ``[\hspace{1em}] how [ ] and pleasing is the form of her face. And how
    \versenum{3}
    soft and how fine is the hair of her head!
    How beautiful are her eyes and how desirable her nose and the whole blossom 
    \versenum{4}
    of her face! [\hspace{1em}] how beautiful are her breasts and how pleasing is all her fairness! Her arms? How beautiful! Her hands? How 
    \versenum{5}
    perfect and desirable is every sight of her hands! How pleasing  are her palms and how straight and petite are all the digits of her hands! Her feet?
    \versenum{6}
    How pleasing! How perfect are her thighs! No virgin or bride who enters into the bridal chamber is more beautiful than her. Surpassing all
    \versenum{7}
    women's beauty is her beauty, and her beauty is above all of them. And with all this beauty, great wisdom is with her, and that which she possess is 
    \versenum{8}
    beautiful.'' And when the king heard the words of Hyrcanus and the words of his two companions (the three of whom spoke as one), he greatly desired her. So he sent
    \versenum{9}
    to quickly lead her (to him). And he looked and he was amazed on account of all her beauty. And he took her as a wife and he sought to kill me, but Sarai said
    \versenum{10}
    to the king, ``he is my brother.'' Therefore I profited because of her and I was left alone---I Abram---and I was not killed. But I wept
    \versenum{11}
    ---I Abram---grievously---both me and Lot, my brother's son with me---in the night when Sarai was taken from me by force.
\end{translation}

\begin{aramaictext}
    \versenum{12}
    בליליא דן צלית ובעית ואתחננת ואמרת באתעצבא ודמעי נחתן בריך אנתה
    אל עליון מרי לכול
    \versenum{13}
    עלמים די אנתה מרה ושליט על כולא ובכול מלכי ארעא אנתה שליט
    למעבד בכולהון דין וכען
    \versenum{14}
    קבלתך מרי על פרעו צען מלך מצרין די דברת אנתתי מני בתוקף עבד לי
    דין מנה ואחזי ידך רבתא
    \versenum{15}
    בה ובכול ביתה ואל ישלט בליליא דן לטמיא אנתתי מני וי\textsuperscript{נ}דעוך מרי די
    אנתה מרה לכול מלכי
    \versenum{16}
    ארעא ובכית וחשית בליליא דן שלח לה אל עליון רוח מכדש למכתשה
    ולכול אנש ביתה רוח
    \versenum{17}
    באישא והואת כתשא לה ולכול אנש ביתה ולא יכל למקרב בהא ואף לא
    ידעהא והוא עמה
    \versenum{18}
    תרתין שנין ולסוף תרתין שנין תקפו וגברו עלוהי מכתשיא ונגדיא ועל כול
    אנש ביתה ושלח
    \versenum{19}
    קרא לכול ח֯כ֯י֯מ̇[י] מצרין ולכול אשפיא עם כול אסי מצרין הן יכולון
    לאסי֯ו֯תה מן מכתשה דן ולאנש
    \versenum{20}
    ביתה ול֯א֯ י֯כ֯לו֯ כ֯ול א̇ס̇יא֯ ואשפיא וכול חכימיא למקם לאסיותה ארי הוא
    רוחא כתש לכולהון
    \versenum{21}
    וערקו \vacat
\end{aramaictext}
\begin{translation}
    \versenum{12}
    That night I prayed and begged and sought mercy in sadness---my tears running down, (saying) ``Blessed are you, God Most High, the Lord of all
    \versenum{13}
    eternity, because you are the Lord and ruler over everything and over every king of the earth you have the power to make judgment on all of them. And now,
    \versenum{14}
    I petition you, oh Lord against Pharoah Zoan, the king of Egypt, because my wife was taken from me by force. Make a restitution for me from him and make visible your mighty hand 
    \versenum{15}
    upon him and upon his whole house. Let him not have the power this night to defile my wife from me. And they will know you, my Lord, that you are the Lord of all the kings of
    \versenum{16}
    the earth.'' And I wept silently. That night, God Most High sent him a spirit of disease to strike him and all the men of his house---an evil
    \versenum{17}
    spirit---and it tormented him and all the men of his house such that he was not able to to come near to her; indeed he did not know her though she was with him 
    \versenum{18}
    for two years. And at the end of two years the plagues and scourges became stronger and prevailed over them and over all the men of his house. So he sent out 
    \versenum{19}
    a call for all the wise men of Egypt and all the magicians with all the physicians of Egypt (to see) if they would be able to heal him from the plague and for the men of 
    \versenum{20}
    his house. But all the physicians and magicians and all the wise men were not able to heal him.  Instead, the spirit afflicted all of them
    \versenum{21}
    and they ran away.
\end{translation}

\begin{aramaictext}
    באדין אתה עלי חרקנוש ובעא מני די אתה ואצלה על
    \versenum{22}
    מלכא֯ ו֯אסמ֯וך ידי עלוהי ויחה ארי ב[ח]לם ח֯ז[ני] ואמר֯ לה לו֯ט לא יכ֯ו֯ל
    אברם דדי לצלי֯א על
    \versenum{23}
    מלכא ושרי אנ֯ת֯ת֯ה֯ ע̇מ̇ה וכען אזל אמר למלכא וישלח אנתתה מנה לבעלהא
    ויצלה עלוהו ויחה
\end{aramaictext}

\begin{translation}
    Then Hyrcanus came to me and and asked that I might pray on behalf of
    \versenum{22}
    the king and that I might put my hands upon him that he might live because [he saw me in a dream]. And Lot said to him, ``Abram, my uncle is not able to pray for
    \versenum{23}
    the king while Sarai his wife is with him. Now, depart and tell the king that he should send his wife from him to her husband so that he will pray for him and live.''
\end{translation}

\begin{aramaictext}
    \versenum{24}
    \vacat
    ו֯כ֯די שמע חרקנוש מלי לוט אזל אמר למלכא כול מכתשיא ונגדיא
    \versenum{25}
    אלן די מתכתש ומתנגד מר̇י מלכא בדיל שרי אנתת אברם י֯ת֯יבו נה לשרי
    לא̇ב֯רם בעלה
    \versenum{26}
    ויתוך מ̇נכה מ̇כ̇ת̇ש̇א̇ דן ורוח שחלניא וקרא [מ]ל[כ]א לי ואמר לי מ̇א עבדתה
    לי בדיל [שר]י ותאמר
    \versenum{27}
    לי די אחתי היא והיא הואת אנתתך ונסבתהא לי לאנתה הא אנתתך ד̇ב֯ר̇ה֯
    א̇זל ועדי לך מן
    \versenum{28}
    כול מדינת מצרין וכען צ֯לי ע֯לי ו֯ע֯ל ביתי ותתגער מננה רוחא דא באיש֯ת̇א
    וצ̇לית֯ עלו֯ה֯י֯ מ֯ג֯דפא
    \versenum{29}
    הו וסמכת ידי ע֯ל֯ [ראי]שה ואתפלי מנה מכתשא ואתגערת [מנה רוחא]
    ב֯אישתא וחי ו֯ק֯ם ו֯י֯הב
    \versenum{30}
    ל̇י֯ מלכא ב[יומא] ד̇נא̇ מנתנ[ן] ש̇גיא̇ן וימ֯א לי מלכא במומה די לא֯ ◦◦◦[  
    ]הא ואת̇יב̇ לי
    \versenum{31}
    לש̇ר̇י ויהב לה מלכא̇ [כסף וד]הב ש̇גיא ולבוש שגי די בוץ וארגואן ו[\hspace{1em}]
    \versenum{32}
    קודמיהא ואף להגר וא[ש]למה לי ומני עמי אנוש די ינפקו֯נני֯ ול◦◦◦ מן
    מ̇צ̇רין \vacat
\end{aramaictext}

\begin{translation}
    \versenum{24}
    When Hyrcanus heard  the words of Lot, he departed and said to the king, ``All these scourges and plagues
    \versenum{25}
    that have scourged and plagued my lord the king  are on account of Sarai, the wife of Abram. Let him return Sarai to Abram, her husband,
    \versenum{26}
    so that this plague might cease from (afflicting) you and also the spirit of oozing.'' And the king summoned me and he said to me, ``What have you done to me on account of [Sarai]? You said
    \versenum{27}
    to me `she is my sister,' but she is really your wife! I took her for myself as a wife! Here is your wife; take her, depart, and go from
    \versenum{28}
    all he provinces of Egypt. But now, pray for me and for my house that this evil spirit be exorcised from us.'' And I prayed for him, that blasphemer,
    \versenum{29}
    and I put my hands upon his [head] and the plague was cleansed from him and the evil spirit was driven from him and he lived. And the king stood up and gave
    \versenum{30}
    to me on that [day] many gifts. And the king swore an oath to me that [he had not defiled her?]. And he returned Sarai to me
    \versenum{31}
    and the king gave her much [silver and] gold and lots of linen and purple linen and [\hspace{1em}]
    \versenum{32}
    before her and also Hagar and he returned her to me and allotted men for me who brought me out and [\hspace{0.5em}] from Egypt.
\end{translation}